---
title: "Estimand–Estimator Mismatch in Time-to-Event Safety Analyses"
subtitle: "A simulation study using longitudinal targeted learning"
format: html
---

<!--
TODO(Joy): verify variable mapping from DGP.R to manuscript notation.
TODO(Joy): refine hepatitis B DGP parameters against published nephrotoxicity rates.
TODO(Joy): decide whether to retain the time-dependent Cox model.
-->

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  fig.align = "center",
  cache = TRUE,
  echo = FALSE,
  message = FALSE,
  warning = FALSE
)
```

```{r}
#| label: load-packages
library(tidyverse)
library(survival)
library(here)
library(knitr)
```

```{r}
#| label: source-helpers
source(here("DGP.R"))
source(here("R", "helpers.R"))
source(here("R", "support_diagnostics.R"))

cache_files <- c(
  here("results", "lmtp_main.rds"),
  here("results", "support_estimation.rds"),
  here("results", "sim_study_main.rds"),
  here("results", "sim_results", "ground_truth.rds")
)
missing <- cache_files[!file.exists(cache_files)]
if (length(missing) > 0) {
  warning("Missing cached results. Run R/run_cached_analyses.R first:\n",
          paste(" ", basename(missing), collapse = "\n"))
}
```

## Introduction

Post-marketing safety analyses routinely ask whether a drug increases the risk of an adverse event. When treatment switching occurs during follow-up---and when switching is driven by evolving clinical indicators that also affect the outcome---the answer depends on how switching is handled. The ICH E9(R1) addendum (2019) formalises this dependence through the estimand framework, requiring that each analysis specify the population, treatment, outcome, intercurrent events, and the strategy used for each intercurrent event.

Different strategies define different causal quantities. A treatment-policy strategy follows all patients from initiation regardless of switching; a hypothetical strategy models a world where switching does not occur; a while-on-treatment strategy restricts attention to the period before switching. Each question has clinical relevance, and each requires a compatible estimator.

In practice, analysts frequently apply Cox regression without verifying that the model's target---a conditional hazard ratio---corresponds to the intended estimand. This paper examines the consequences of that mismatch using a simulation study built around a hepatitis B safety scenario. We compare conventional Cox analyses to estimation via the longitudinal modified treatment policy (LMTP) framework, which directly targets marginal risk contrasts under specified interventions. The analysis concentrates on two estimands---treatment-policy and hypothetical no-switch---and illustrates how the same data yield different estimates depending on the strategy employed and the degree to which the estimator matches it.

We follow the causal roadmap for real-world evidence (Dang et al. 2023; Petersen 2014), concentrating on estimand definition, identifiability assessment, and estimator selection.

## Motivating Setting

The simulation draws on concerns about nephrotoxicity in tenofovir-containing hepatitis B regimens relative to entecavir-based alternatives. Treatment assignment in the simulated cohort is confounded by baseline CKD, cirrhosis, and age. The primary intercurrent event is informative treatment switching: patients on the active regimen who develop early renal signals may switch to the comparator. Because CKD drives both switching and renal failure, switching is informative.

<!-- TODO(Joy): add a paragraph with real-data references from the
     hepatitis B nephrotoxicity literature if available. -->

## Simulation Design

### Data-generating process

The function `generate_hep_data()` produces cohorts with confounded treatment assignment, a piecewise-exponential event process with non-proportional hazards (HR = 1.25 before day 45; HR = 0.70 after), and informative switching driven by treatment and CKD. When a patient switches, the post-switch event hazard changes to reflect the new treatment---that is, switching has a direct effect on the outcome process. Administrative censoring may depend on baseline risk when `dep_censor = TRUE`.

The DGP supports several follow-up policies. Under `"treatment_policy"`, switching occurs and modifies the hazard, but does not censor follow-up. Under `"no_switch"`, switching is suppressed entirely. Under `"while_on_treatment"`, follow-up is censored at the switch time.

```{r}
#| label: generate-data
#| cache: true
set.seed(2026)
dat <- generate_hep_data(
  N           = 10000,
  np_hazard   = TRUE,
  dep_censor  = TRUE,
  complexity  = TRUE,
  policy      = "treatment_policy",
  seed        = 2026
)
```

```{r}
#| label: data-summary

data_summary <- tibble::tibble(
  Quantity = c("Sample size", "180-day event rate", "Switching rate",
               "Switching rate (active, CKD)", "Switching rate (comparator)",
               "Median follow-up (days)"),
  Value = c(format(nrow(dat), big.mark = ","),
            round(mean(dat$event), 3),
            round(mean(dat$switched), 3),
            round(mean(dat$switched[dat$treatment == 1 & dat$ckd == 1]), 3),
            round(mean(dat$switched[dat$treatment == 0]), 3),
            round(median(dat$follow_time), 0))
)
kable(data_summary, format = "pipe", caption = "Simulated cohort characteristics")
```

```{r}
#| label: fig-dag
#| fig-cap: "Assumed causal structure. Baseline confounders (W) affect treatment assignment, switching, censoring, and renal failure. Switching modifies the post-switch hazard for renal failure."
#| fig-width: 8
#| fig-height: 5
library(dagitty)
library(ggdag)

dag_safety <- dagify(
  Treatment ~ W,
  Switch ~ Treatment + W,
  Censoring ~ Switch + W,
  RenalFailure ~ Treatment + Switch + W,
  exposure = "Treatment", outcome = "RenalFailure",
  labels = c(W = "Baseline\nconfounders\n(CKD, age, ...)",
             Treatment = "Treatment\n(active vs comparator)",
             Switch = "Regimen\nswitch",
             Censoring = "Censoring",
             RenalFailure = "Renal\nfailure"),
  coords = list(
    x = c(W = 1, Treatment = 0, Switch = 1, Censoring = 2, RenalFailure = 1.5),
    y = c(W = 1, Treatment = 0, Switch = 0, Censoring = 0, RenalFailure = -1)
  )
)
ggdag(dag_safety, text = FALSE, use_labels = "label", text_size = 3) +
  theme_dag()
```

### Known truth

Because the DGP is fully specified, true marginal risks can be computed by generating large counterfactual datasets (N = 500,000 per arm) under each policy, with no dependent censoring.

```{r}
#| label: compute-truth

truth_cache <- here("results", "sim_results", "ground_truth.rds")
stopifnot("Run R/run_cached_analyses.R first" = file.exists(truth_cache))
truth_all <- readRDS(truth_cache)
```

```{r}
#| label: tbl-truth
#| tbl-cap: "True 180-day risks by estimand strategy (Monte Carlo, N = 500,000 per arm)"

truth_summary <- tibble::tibble(
  Strategy = c("Treatment-policy", "Hypothetical no-switch",
               "While-on-treatment", "Composite", "Principal stratum"),
  `Risk (active)` = c(truth_all$treatment_policy$true_risk_1,
                       truth_all$no_switch$true_risk_1,
                       truth_all$while_on_treatment$true_risk_1,
                       truth_all$composite$true_risk_1,
                       truth_all$principal_stratum$true_risk_1),
  `Risk (comparator)` = c(truth_all$treatment_policy$true_risk_0,
                           truth_all$no_switch$true_risk_0,
                           truth_all$while_on_treatment$true_risk_0,
                           truth_all$composite$true_risk_0,
                           truth_all$principal_stratum$true_risk_0),
  `Risk difference` = c(truth_all$treatment_policy$true_rd,
                         truth_all$no_switch$true_rd,
                         truth_all$while_on_treatment$true_rd,
                         truth_all$composite$true_rd,
                         truth_all$principal_stratum$true_rd),
  `Marginal HR` = c(truth_all$treatment_policy$true_hr,
                     truth_all$no_switch$true_hr,
                     truth_all$while_on_treatment$true_hr,
                     truth_all$composite$true_hr,
                     truth_all$principal_stratum$true_hr)
)
kable(truth_summary, digits = 4, format = "pipe")
```

The treatment-policy risk difference is positive: the active drug causes early harm (HR = 1.25 before day 45), and although switching to the comparator reduces the subsequent hazard, the accumulated early risk is not fully offset. Under the no-switch strategy, where patients remain on their assigned treatment for the full 180 days, the early harm and late protective effect (HR = 0.70 after day 45) approximately cancel. The composite strategy produces a much larger risk difference because the active arm experiences substantially more switching, and switching itself counts as an event. The principal stratum restricts to the approximately 19% of patients who would not switch under either treatment; this low-risk subgroup shows a near-null treatment effect.

The while-on-treatment truth is numerically identical to the no-switch truth in this design. Under counterfactual uniform treatment (all-treated or all-control), no switching occurs, so censoring at switch has no effect. In observed data where both treatments are assigned, the two estimands would differ: while-on-treatment conditions on not having switched, whereas the hypothetical intervenes to prevent switching.

## Estimand Framework

The ICH E9(R1) addendum defines five strategies for handling intercurrent events. We briefly state each in the context of treatment switching, then concentrate on the two that are most directly supported by the current simulation.

**Treatment-policy.** Compare 180-day risks between populations initiated on active versus comparator, regardless of subsequent switching. The target is $\psi^{TP} = E[Y^{a=1}(180)] - E[Y^{a=0}(180)]$, where switching occurs naturally.

**Hypothetical no-switch.** Compare risks under the counterfactual assumption that no patients switch. The target is $\psi^{HYP} = E[Y^{a=1, \text{no switch}}(180)] - E[Y^{a=0, \text{no switch}}(180)]$. Identification requires modelling the confounders that drive switching.

**While-on-treatment.** Compare risks while patients remain on assigned treatment. Follow-up is censored at switching. Naive estimation assumes non-informative censoring; IPCW or LMTP can relax this.

**Composite.** Define the outcome as renal failure *or* switching. This avoids modelling switching but conflates tolerability with safety.

**Principal stratum.** Restrict to patients who would not switch under either treatment. The stratum is latent and requires structural assumptions for identification.

The main analyses below focus on the treatment-policy and no-switch strategies, which illustrate the estimand--estimator mismatch most directly. Composite and principal stratum results from the simulation are reported in supplementary tables but are not discussed in detail because the current LMTP implementation does not provide well-matched estimators for them (see Limitations).

## Estimation Strategies

### Cox regression

Three Cox-based approaches are commonly used in the presence of switching.

```{r}
#| label: cox-fits
tau <- 180
covars <- c("age", "ckd", "cirrhosis", "diabetes", "heart_failure")
cox_naive <- fit_cox_naive(dat, tau = tau, covars = covars)
cox_cs    <- fit_cox_censor_switch(dat, tau = tau, covars = covars)
cox_td    <- fit_cox_td(dat, tau = tau, covars = covars)
km_naive  <- km_risk_at(cox_naive$survfit, t = tau)
km_cs     <- km_risk_at(cox_cs$survfit, t = tau)
```

```{r}
#| label: tbl-cox
#| tbl-cap: "Cox regression estimates (single dataset, N = 10,000). KM risk difference is unadjusted."

cox_tbl <- tibble::tibble(
  Analysis = c("Baseline treatment only",
               "Censor at switch",
               "Time-dependent treatment"),
  `Adjusted HR` = round(c(cox_naive$hr, cox_cs$hr, cox_td$hr), 3),
  `HR 95% CI` = c(
    paste0(round(cox_naive$ci_low, 3), "--", round(cox_naive$ci_high, 3)),
    paste0(round(cox_cs$ci_low, 3), "--", round(cox_cs$ci_high, 3)),
    paste0(round(cox_td$ci_low, 3), "--", round(cox_td$ci_high, 3))),
  `KM risk diff.` = c(round(km_naive$risk_diff, 4),
                       round(km_cs$risk_diff, 4), NA)
)
kable(cox_tbl, format = "pipe")
```

The baseline-treatment Cox model ignores switching and is often interpreted as a treatment-policy analysis. Under non-proportional hazards, however, the Cox HR is a weighted average of time-specific hazard ratios, and the weights depend on the censoring distribution (Rufibach 2018). When CKD confounds both treatment and the outcome, the unadjusted KM risk difference reflects confounding rather than a causal contrast.

The censor-at-switch model is sometimes used to approximate a hypothetical or while-on-treatment analysis. It assumes that switching is non-informative given baseline covariates. In this DGP, CKD drives both switching and renal failure, so the assumption is violated: censoring selectively removes high-risk person-time from the active arm.

The time-dependent model includes current treatment status as a covariate and estimates a conditional HR for current exposure. This does not correspond to a marginal risk contrast under any of the estimand strategies above, and requires the strong assumption of no unmeasured time-varying confounding of the switching decision.

Each Cox analysis estimates a conditional hazard ratio. None directly targets the marginal risk difference defined by either the treatment-policy or hypothetical estimand. This distinction---between the parameter a model estimates and the parameter the scientific question requires---is the central concern of this paper.

::: {.callout-note}
## On the hazard ratio as a causal parameter
Even in a randomised trial, the population-level HR at time $t$ compares groups whose composition has been differentially altered by prior events (Stensrud et al. 2019). The individual-level causal HR requires conditional independence of counterfactual survival times (Martinussen 2022), an untestable assumption. Under non-proportional hazards, the Cox HR additionally depends on the censoring distribution. Survival-function contrasts---risk differences, risk ratios, restricted mean survival time---avoid several of these difficulties.
:::

### LMTP

The `lmtp` package (Williams and Diaz 2023) implements longitudinal modified treatment policies via the sequentially doubly robust (SDR) estimator. It targets a marginal survival probability under a specified intervention, adjusting for time-varying confounding, and provides inference that is consistent if either the outcome model or the treatment/censoring model is correctly specified.

```{r}
#| label: lmtp-estimates
lmtp_cache <- here("results", "lmtp_main.rds")
stopifnot("Run R/run_cached_analyses.R first" = file.exists(lmtp_cache))
lmtp_res <- readRDS(lmtp_cache)
rd <- lmtp_res$contrast_rd$estimates
rr <- lmtp_res$contrast_rr$estimates
```

```{r}
#| label: tbl-lmtp
#| tbl-cap: "LMTP estimates under static interventions (single dataset, N = 10,000, biweekly bins)"

lmtp_tbl <- tibble::tibble(
  Contrast = c("Risk difference (active − comparator)",
               "Risk ratio (active / comparator)"),
  Estimate = c(round(rd$estimate, 4), round(rr$estimate, 3)),
  `95% CI` = c(
    paste0(round(rd$conf.low, 4), " to ", round(rd$conf.high, 4)),
    paste0(round(rr$conf.low, 3), " to ", round(rr$conf.high, 3)))
)
kable(lmtp_tbl, format = "pipe")
```

The LMTP analysis sets baseline treatment to active ($A = 1$) or comparator ($A = 0$) for all subjects and estimates marginal 180-day risks under each intervention. Follow-up is discretised into biweekly bins (13 time points). The current Super Learner library includes `SL.mean`, `SL.glm`, and `SL.bayesglm` with 2-fold cross-validation; production analyses should use a richer library and more folds (Gruber et al. 2022).

Two important qualifications apply. First, the LMTP analysis as implemented here uses static baseline interventions and a censoring model that adjusts for informative censoring (including switching-induced follow-up loss). It provides a marginal risk contrast that is better aligned with the treatment-policy estimand than any of the Cox approaches, but the degree of alignment depends on correct specification of the censoring and outcome models. Second, a fully distinct LMTP analysis for the hypothetical no-switch estimand would require an intervention function that sets treatment at all post-baseline time points to the baseline value, preventing switching. The current implementation uses the same static baseline intervention for both strategies, and the two estimands are distinguished by generating data under different DGP policies rather than by different LMTP intervention specifications. This is an acknowledged limitation.

## Empirical Support Diagnostics {#sec-support}

Positivity requires that the probability of receiving each treatment remains bounded away from zero within every confounder stratum. In longitudinal settings, this extends to a sequential condition on treatment and censoring probabilities at each time point (Gruber et al. 2022). We examine three scenarios that vary switching intensity while holding treatment assignment fixed.

```{r}
#| label: support-scenarios
#| cache: true
set.seed(101)
dat_good <- generate_hep_data(
  N = 5000, np_hazard = TRUE, dep_censor = TRUE,
  complexity = TRUE, policy = "treatment_policy", seed = 101)
set.seed(102)
dat_strained <- generate_hep_data(
  N = 5000, np_hazard = TRUE, dep_censor = TRUE,
  complexity = TRUE, policy = "treatment_policy",
  gamma_A = 1.5, gamma_ckd = 1.2, lambda_sw0 = 5e-3, seed = 102)
set.seed(103)
dat_poor <- generate_hep_data(
  N = 5000, np_hazard = TRUE, dep_censor = TRUE,
  complexity = TRUE, policy = "treatment_policy",
  gamma_A = 2.5, gamma_ckd = 2.0, lambda_sw0 = 1e-2, seed = 103)
```

Because treatment assignment is identical across scenarios, treatment propensity scores do not change. The relevant diagnostic is the predicted switching probability, which shifts substantially as switching intensity increases.

```{r}
#| label: fig-support
#| fig-cap: "Predicted switching probability by treatment arm across three support scenarios."
#| fig-height: 9
#| fig-width: 7
scenario_list <- list(
  "1. Good support" = dat_good,
  "2. Strained support" = dat_strained,
  "3. Poor support" = dat_poor)
support_plots <- plot_support_scenarios(scenario_list)
support_plots$by_treatment
```

```{r}
#| label: fig-support-ckd
#| fig-cap: "Switching probability by CKD status. Under poor support, CKD patients switch near-deterministically."
#| fig-height: 9
#| fig-width: 7
support_plots$by_ckd
```

```{r}
#| label: tbl-support-est
#| tbl-cap: "Estimation across support scenarios"
support_cache <- here("results", "support_estimation.rds")
stopifnot("Run R/run_cached_analyses.R first" = file.exists(support_cache))
support_results <- readRDS(support_cache)
kable(support_results, digits = 3, format = "pipe")
```

Under good support (~12% switching), all estimators produce stable results. As switching intensity increases, the censor-at-switch HR diverges from the baseline-treatment HR because censoring removes a progressively larger and more selected subset of the active arm. LMTP estimates show wider confidence intervals and increased bias as the censoring model must extrapolate into poorly supported regions.

## Simulation Results

We assess estimator performance over 20 repeated samples (N = 10,000 each) across all five DGP policies. For each iteration we fit baseline-treatment Cox, censor-at-switch Cox, and LMTP, then compare the estimated risk differences to the policy-specific truth.

```{r}
#| label: sim-load
source(here("R", "run_simulations.R"))
sim_cache <- here("results", "sim_study_main.rds")
stopifnot("Run R/run_cached_analyses.R first" = file.exists(sim_cache))
sim_results <- readRDS(sim_cache)
```

The primary comparison focuses on the treatment-policy and no-switch policies. Results for the remaining three policies are included for completeness but should be interpreted with caution: the composite LMTP analysis is known to be misspecified (the censoring model adjusts away switching, whereas the composite estimand counts switching as an event), and the principal stratum analysis uses an approximation (restricting to observed non-switchers rather than the latent never-switcher stratum).

```{r}
#| label: tbl-sim-main
#| tbl-cap: "Simulation results for treatment-policy and no-switch estimands (20 iterations, N = 10,000). Bias and coverage are computed relative to the policy-specific truth."

estimand_labels <- c(
  treatment_policy   = "Treatment-policy",
  no_switch          = "No-switch",
  while_on_treatment = "While-on-treatment",
  composite          = "Composite",
  principal_stratum  = "Principal stratum"
)

sim_summaries <- lapply(names(truth_all), function(est) {
  if (!est %in% sim_results$estimand) return(NULL)
  sub <- sim_results %>% filter(estimand == est)
  s <- summarize_simulation(sub,
         truth_rd = truth_all[[est]]$true_rd,
         truth_rr = truth_all[[est]]$true_rr,
         truth_hr = truth_all[[est]]$true_hr)
  s$estimand_label <- estimand_labels[est]
  s
}) %>% bind_rows()

# Main table: treatment-policy and no-switch only
sim_main <- sim_summaries %>%
  filter(estimand %in% c("treatment_policy", "no_switch")) %>%
  select(estimand_label, method, n_converged,
         mean_hr, bias_hr, mean_rd, bias_rd, coverage_rd) %>%
  rename(Strategy = estimand_label, Estimator = method,
         `n conv.` = n_converged,
         `Mean HR` = mean_hr, `HR bias` = bias_hr,
         `Mean RD` = mean_rd, `RD bias` = bias_rd,
         `RD coverage` = coverage_rd) %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

kable(sim_main, digits = 4, format = "pipe")
```

```{r}
#| label: fig-sim-rd
#| fig-cap: "Distribution of risk difference estimates across 20 iterations for the treatment-policy and no-switch strategies. Dashed line: true RD."
#| fig-height: 5
#| fig-width: 9

sim_plot_main <- sim_results %>%
  filter(converged, !is.na(risk_diff),
         estimand %in% c("treatment_policy", "no_switch")) %>%
  mutate(estimand_label = estimand_labels[estimand])

truth_main <- tibble(
  estimand_label = c("Treatment-policy", "No-switch"),
  xint = c(truth_all$treatment_policy$true_rd,
           truth_all$no_switch$true_rd))

ggplot(sim_plot_main, aes(x = risk_diff, fill = method)) +
  geom_histogram(alpha = 0.6, bins = 15, position = "identity") +
  facet_wrap(~ estimand_label, scales = "free_x") +
  geom_vline(data = truth_main,
             aes(xintercept = xint), linetype = "dashed", color = "red") +
  labs(x = "Risk difference", y = "Count", fill = "Estimator") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "bottom")
```

### Supplementary: remaining strategies

```{r}
#| label: tbl-sim-supp
#| tbl-cap: "Simulation results for while-on-treatment, composite, and principal stratum strategies (20 iterations). These results should be interpreted with caution; see Limitations."

sim_supp <- sim_summaries %>%
  filter(!estimand %in% c("treatment_policy", "no_switch")) %>%
  select(estimand_label, method, n_converged,
         mean_hr, bias_hr, mean_rd, bias_rd, coverage_rd) %>%
  rename(Strategy = estimand_label, Estimator = method,
         `n conv.` = n_converged,
         `Mean HR` = mean_hr, `HR bias` = bias_hr,
         `Mean RD` = mean_rd, `RD bias` = bias_rd,
         `RD coverage` = coverage_rd) %>%
  mutate(across(where(is.numeric), ~ ifelse(is.nan(.), NA, .)))

kable(sim_supp, digits = 4, format = "pipe")
```

## Discussion

The simulation illustrates two distinct sources of estimand--estimator mismatch in the presence of informative switching.

The first is confounding bias. The baseline-treatment Cox model and unadjusted KM risk differences consistently overestimate the treatment effect across all strategies, with 0% coverage in every case. CKD confounds both treatment assignment and renal failure, and the Cox model's conditional HR does not correspond to the marginal risk contrast defined by either the treatment-policy or hypothetical estimand. This is not a deficiency specific to the Cox model; it reflects the use of an estimator that does not adjust adequately for confounding and does not target the intended parameter.

The second source of mismatch is how switching is handled. The censor-at-switch Cox analysis selectively removes high-risk person-time, attenuating the apparent treatment effect relative to the baseline-treatment model. The degree of attenuation increases with switching intensity: under poor support (~67% switching), the censor-at-switch HR diverges substantially from the naive HR. Neither Cox analysis distinguishes between the treatment-policy and no-switch estimands in a principled way; both estimate conditional hazard ratios that do not correspond to either marginal target.

LMTP estimates are less biased than the Cox analyses for the treatment-policy and no-switch strategies, with coverage reaching 75--85% for the near-null no-switch truth. Coverage for the treatment-policy strategy is lower (approximately 35%), reflecting residual bias that likely arises from the combination of coarse time discretisation (biweekly bins), a minimal Super Learner library, and a sparse event rate. The LMTP analysis provides a marginal risk contrast that is better aligned with the intervention-defined estimand than any of the Cox approaches, but it does not fully recover the truth under the current specification.

### Sensitivity to unmeasured confounding

Causal interpretation of any observational estimate depends on the assumption of no unmeasured confounding. The causal gap $\delta = \psi_{\text{stat}} - \psi_{\text{causal}}$ quantifies the discrepancy between the statistical and causal estimands. The G-value (Gruber et al. 2024)---the smallest $\delta$ that would shift the confidence interval to include the null---provides a useful diagnostic. A full sensitivity analysis is not implemented here but would be a necessary component of any production analysis.

## Limitations

The LMTP analysis for the no-switch and while-on-treatment estimands models treatment as time-varying and applies a static intervention that holds treatment constant across all time bins. This directly implements the "no switching" intervention within LMTP. For the treatment-policy estimand, treatment is modelled as time-fixed (baseline only), and the LMTP censoring model handles informative follow-up loss due to switching. The degree of alignment depends on correct specification of the censoring and outcome models.

The while-on-treatment truth is computed as the conditional risk among subjects who would not switch before the 180-day risk window under counterfactual uniform treatment. This exercises the switching mechanism but remains a conditional quantity (conditioning on a post-treatment variable), which differs conceptually from an interventional target.

The composite LMTP analysis uses the DGP's composite outcome (renal failure or switching, whichever comes first) directly. Under the composite policy, switching is encoded in the event indicator rather than the censoring indicator, so the LMTP censoring model adjusts only for administrative censoring. This addresses the prior misspecification concern, though the composite estimand's clinical interpretability remains limited.

For the principal stratum, two LMTP analyses are reported: one restricted to observed non-switchers (a convenience approximation) and one restricted to true never-switchers identified through both potential switching outcomes. The true-stratum analysis is possible because the DGP generates paired switching indicators under each treatment assignment. In practice, the never-switcher stratum is latent and would require structural assumptions (e.g., monotonicity) or sensitivity analysis for identification.

Biweekly time bins and a minimal Super Learner library (`SL.mean`, `SL.glm`, `SL.bayesglm`) are used throughout. Finer discretisation and a richer learner set would likely reduce residual bias. The simulation uses 20 iterations; coverage estimates at this sample size are imprecise.

<!-- TODO(Joy): increase iterations and enrich SL library for production. -->

## References

- ICH E9(R1) Addendum (2019). Addendum on estimands and sensitivity analyses in clinical trials.
- Williams N, Diaz I (2023). lmtp: Non-parametric causal effects of feasible interventions based on modified treatment policies. R package.
- van der Laan MJ, Rose S (2011). *Targeted Learning.* Springer.
- Hernan MA, Robins JM (2020). *Causal Inference: What If.* Chapman & Hall.
- Dang LE, Tarp JM, et al. (2023). A causal roadmap for generating high-quality real-world evidence. *J Clin Transl Sci.*
- Gruber S, Lee H, Phillips R, Ho M, van der Laan MJ (2022). Developing a targeted learning-based statistical analysis plan. *Stat Biopharm Res.*
- Gruber S, Phillips RV, Lee H, van der Laan MJ (2024). Targeted learning: toward a future informed by real-world evidence. *Stat Biopharm Res.*
- Chen RC, Rosenblum M, van der Laan MJ, Gruber S (2023). Beyond the Cox hazard ratio. *Clinical Trials.*
- Martinussen T (2022). Causality and the Cox regression model. *Annu Rev Stat Appl.*
- Stensrud MJ, Aalen JM, Aalen OO, Valberg M (2019). Limitations of hazard ratios in clinical trials. *Eur Heart J.*
- Rufibach K (2019). Treatment effect quantification for time-to-event endpoints. *Pharm Stat.*
- Petersen ML (2014). Applying a causal road map in settings with time-dependent confounding. *Epidemiology.*
