---
title: "Estimand-Estimator Alignment in Time-to-Event Safety Analyses"
subtitle: "A hepatitis B case study with longitudinal targeted learning"
format: html
---

<!--
TODO(Joy): verify exact variable mapping from DGP.R to chapter notation.
TODO(Joy): refine hepatitis B DGP parameters to match realistic renal-failure
  incidence in tenofovir vs entecavir from the literature.
TODO(Joy): replace placeholder intervention function with the final
  study-specific rule if needed.
TODO(Joy): decide whether to include the time-dependent Cox model in the
  main presentation or relegate it to an appendix.
-->

# Estimand-Estimator Alignment in Time-to-Event Safety Analyses

## Introduction

In real-world safety studies, the question "Does this drug increase the risk of adverse events?" requires a precise definition of the causal quantity being estimated. The ICH E9(R1) addendum formalises this through the **estimand framework**, which requires specifying how intercurrent events---treatment switching, discontinuation, death from other causes---are handled. Each strategy defines a different estimand, and each estimand demands a compatible estimator.

The central problem this chapter addresses is not which estimand to choose. Rather, it is that **conventional regression analyses often fail to target the stated estimand** when intercurrent events are present. A naive Cox model that ignores switching, or one that censors at switching, may appear to estimate a treatment effect, but the quantity it targets can be misaligned with the scientific question. Longitudinal targeted learning via `lmtp` offers a principled alternative that directly targets the intervention-defined estimand and adjusts for time-varying confounding.

This chapter focuses on two estimands---the **treatment-policy** estimand and the **hypothetical no-switch** estimand---and demonstrates how to align (or misalign) estimators with each. The goal is not to compare estimands against each other, but to ensure that the chosen estimator matches the causal question.

This chapter follows the structure of the **causal roadmap** for generating high-quality real-world evidence (Dang et al. 2023; Petersen 2014). The roadmap proceeds through: (1) defining the causal question and estimand as a function of counterfactual outcomes---not as a regression coefficient; (2) specifying a causal model (DAG); (3) assessing identifiability under exchangeability, positivity, and consistency; (4) defining the statistical estimand; (5) choosing an estimator; and (6) conducting sensitivity analyses. This tutorial concentrates on steps 1, 3--5: defining estimands, diagnosing support, and aligning estimators. A key principle of the roadmap is that the estimand should be defined mathematically before any statistical method is chosen, and that the gap between the causal estimand and the statistical estimand should be made explicit.

**Goals:**

1. Define the treatment-policy and hypothetical no-switch estimands under the ICH E9(R1) framework
2. Show how conventional Cox regression analyses can fail to target these estimands
3. Demonstrate how `lmtp`-based longitudinal targeted learning can better align estimation with the stated estimand
4. Diagnose when empirical support for a target intervention is inadequate

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  fig.align = "center",
  cache = TRUE
)
```

```{r}
#| label: load-packages
#| message: false
#| warning: false
library(tidyverse)
library(survival)
library(here)
library(knitr)
```

```{r}
#| label: source-helpers
#| message: false
#| warning: false
source(here("DGP.R"))
source(here("R", "helpers.R"))
source(here("R", "support_diagnostics.R"))

# Check that cached results exist (run R/run_cached_analyses.R first)
cache_files <- c(
  here("results", "lmtp_main.rds"),
  here("results", "support_estimation.rds"),
  here("results", "sim_study_main.rds"),
  here("results", "sim_results", "ground_truth.rds")
)
missing <- cache_files[!file.exists(cache_files)]
if (length(missing) > 0) {
  warning("Missing cached results. Run R/run_cached_analyses.R first:\n",
          paste(" ", basename(missing), collapse = "\n"))
}
```

## The Hepatitis B Safety Example

We use a simulated post-marketing safety study comparing two hepatitis B antiviral regimens for the risk of **renal failure** (the safety endpoint). The setting is motivated by real concerns about nephrotoxicity associated with tenofovir-containing regimens relative to entecavir-based alternatives in patients with chronic hepatitis B.

Key features of the example:

- **Two treatments:** an active regimen (analogous to tenofovir-containing therapy) and a comparator (analogous to entecavir-based therapy), with confounded treatment assignment driven by baseline renal function (CKD), cirrhosis severity, and age.
- **Intercurrent event:** informative treatment switching. Patients on the active regimen who develop early signs of renal impairment may switch to the comparator. Because CKD drives both switching and renal failure, switching is informative.
- **Outcome:** time to renal failure (or a composite renal endpoint), observed over a 180-day risk window.
- **Baseline confounders:** demographics (age, sex, race), comorbidities (CKD, diabetes, hypertension, cirrhosis, heart failure), and co-medications (NSAIDs, ACE/ARBs, statins).

This example illustrates a realistic scenario in which informative switching invalidates naive regression approaches and where the choice of estimand strategy has concrete consequences for the analysis.

<!-- TODO(Joy): add one paragraph with real-data references if a project-approved
     example from the hepatitis B nephrotoxicity literature is available. -->

## The Regulatory Context: ICH E9(R1)

The ICH E9(R1) addendum (2019) requires explicit specification of five elements for any clinical or post-marketing estimand:

1. **The population** of interest
2. **The treatment** conditions being compared
3. **The outcome** variable
4. **The intercurrent events** that may occur after treatment initiation
5. **The strategy** for handling each intercurrent event

When 15--20% of patients on the active regimen switch due to early renal signals, and when patients with CKD are more likely to both switch and develop renal failure, the strategy used to handle switching defines fundamentally different estimands---and different estimands require different estimators.

## The Data-Generating Process

We use `generate_hep_data()` from the project DGP to simulate realistic post-marketing safety data. The DGP produces a cohort of patients eligible for one of two hepatitis B treatments, with renal failure as the safety endpoint.

| DGP Component | Description |
|---|---|
| **Baseline covariates** | Demographics (age, sex, race); comorbidities (CKD, diabetes, hypertension, cirrhosis, heart failure); co-medications (NSAIDs, ACE/ARBs, statins) |
| **Treatment assignment** | Active vs comparator, confounded by CKD, cirrhosis, age, and (when `complexity=TRUE`) nonlinear covariate interactions |
| **Event process** | Piecewise exponential renal-failure hazard; when `np_hazard=TRUE`, HR = 1.25 early and HR = 0.70 late (non-proportional hazards) |
| **Switching process** | Hazard-based informative switching driven by treatment and CKD |
| **Censoring process** | When `dep_censor=TRUE`, censoring depends on baseline risk; otherwise administrative censoring only |

The DGP accepts a `policy` argument:

- `policy = "treatment_policy"`: switching occurs but does **not** censor. Subjects are followed until event or administrative censoring regardless of switching.
- `policy = "no_switch"`: no switching occurs. This generates potential outcomes under the hypothetical world where nobody switches.

<!-- TODO(Joy): verify exact variable mapping from DGP.R to chapter notation -->

```{r}
#| label: generate-data
#| cache: true
set.seed(2026)
dat <- generate_hep_data(
  N           = 5000,
  np_hazard   = TRUE,
  dep_censor  = TRUE,
  complexity  = TRUE,
  policy      = "treatment_policy",
  seed        = 2026
)

cat("N:", nrow(dat), "\n")
cat("Event rate:", round(mean(dat$event), 4), "\n")
cat("Switch rate:", round(mean(dat$switched), 4), "\n")
cat("Switch rate (active + CKD):",
    round(mean(dat$switched[dat$treatment == 1 & dat$ckd == 1]), 4), "\n")
cat("Switch rate (comparator):",
    round(mean(dat$switched[dat$treatment == 0]), 4), "\n")
cat("Median follow-up:", round(median(dat$follow_time), 1), "days\n")
```

Switching is driven by both treatment assignment and CKD status. Patients on the active regimen with CKD switch at the highest rate. Because CKD also increases renal failure risk, switching is **informative**: the patients who switch are those at greatest risk for the outcome.

```{r}
#| label: fig-dag
#| fig-cap: "Causal structure: baseline confounders (W) affect treatment, switching, censoring, and renal failure. Switching is informative because W (especially CKD) drives both switching and the outcome."
#| fig-width: 8
#| fig-height: 5
library(dagitty)
library(ggdag)

dag_safety <- dagify(
  Treatment ~ W,
  Switch ~ Treatment + W,
  Censoring ~ Switch + W,
  RenalFailure ~ Treatment + Switch + W,
  exposure = "Treatment", outcome = "RenalFailure",
  labels = c(W = "Baseline\nconfounders\n(CKD, age, ...)",
             Treatment = "Treatment\n(active vs comparator)",
             Switch = "Regimen\nswitch",
             Censoring = "Censoring",
             RenalFailure = "Renal\nfailure"),
  coords = list(
    x = c(W = 1, Treatment = 0, Switch = 1, Censoring = 2, RenalFailure = 1.5),
    y = c(W = 1, Treatment = 0, Switch = 0, Censoring = 0, RenalFailure = -1)
  )
)
ggdag(dag_safety, text = FALSE, use_labels = "label", text_size = 3) +
  theme_dag()
```

## Two Estimand Strategies

We focus on two ICH E9(R1) strategies that most clearly illustrate the alignment problem. Other strategies (while-on-treatment, composite, principal stratum) exist but are not the focus here.

### Estimand 1: Treatment-Policy

**Question:** What is the risk of renal failure by 180 days comparing populations *initiated* on the active regimen versus the comparator, regardless of subsequent treatment changes?

**Intercurrent event strategy:** Ignore switching. Follow all patients from initiation regardless of what happens to their treatment.

**Target parameter:** $\psi^{TP} = E[Y^{a=1}(180)] - E[Y^{a=0}(180)]$, where $Y^a(t)$ is the counterfactual renal-failure indicator under baseline initiation of regime $a$, with all subsequent intercurrent events occurring as they naturally would.

### Estimand 2: Hypothetical No-Switch

**Question:** What *would* the risk of renal failure be by 180 days if, contrary to fact, no patients switched regimens?

**Intercurrent event strategy:** Model the hypothetical scenario in which switching does not occur. This requires identifying and adjusting for the time-varying confounders that drive both switching and the outcome.

**Target parameter:** $\psi^{HYP} = E[Y^{a=1, \text{no switch}}(180)] - E[Y^{a=0, \text{no switch}}(180)]$

### Why These Two

The treatment-policy estimand is relevant when the clinical question concerns the real-world consequences of initiating a regimen, including all downstream treatment decisions. The hypothetical estimand is relevant when the question concerns the direct pharmacological effect of the drug, absent any treatment modifications. Both are commonly targeted in safety analyses, and both are susceptible to estimand-estimator misalignment when conventional regression is used.

## Estimand-Estimator Alignment

The table below is the conceptual core of this chapter. For each estimand, it lists the aligned LMTP estimator and the misaligned Cox estimators, together with the reason for misalignment.

```{r}
#| label: tbl-alignment
#| tbl-cap: "Estimand versus estimator: aligned and misaligned approaches"

alignment_tbl <- tibble::tribble(
  ~Estimand, ~Estimator, ~Aligned, ~`Reason for misalignment`,
  "Treatment-policy", "LMTP SDR (static on/off)", "Yes",
    "Targets marginal risk under baseline initiation; adjusts for censoring",
  "Treatment-policy", "Cox naive (ignore switch)", "No",
    "Targets conditional HR, not marginal RD; risk set altered by informative switching",
  "Treatment-policy", "Cox censor-at-switch", "No",
    "Censoring at switch removes high-risk subjects; wrong estimand entirely",
  "Treatment-policy", "Cox time-dependent trt", "No",
    "Conditional HR with time-varying exposure; does not target marginal risk",
  "Hypothetical no-switch", "LMTP SDR (no-switch intervention)", "Yes",
    "Targets marginal risk in a world without switching; adjusts for confounding",
  "Hypothetical no-switch", "Cox naive (ignore switch)", "No",
    "Includes post-switch outcomes; does not model the hypothetical world",
  "Hypothetical no-switch", "Cox censor-at-switch", "No",
    "Assumes non-informative censoring; biased when switching is informative"
)

kable(alignment_tbl, format = "pipe")
```

A brief conceptual explanation: when switching is informative (driven by CKD and treatment), censoring at switch removes the highest-risk person-time from the active treatment arm. This makes the active treatment appear safer than it is. The naive Cox model, by contrast, includes post-switch outcomes that may reflect the comparator treatment rather than the active one. Neither targets the marginal risk contrast that defines the estimand. LMTP avoids these problems by directly targeting the counterfactual risk under a specified intervention and adjusting for confounding and censoring through its doubly robust longitudinal estimation.

## Conventional Regression Analyses (Misaligned)

### A. Naive Baseline Cox Model

```{r}
#| label: cox-naive
tau <- 180

cox_naive <- fit_cox_naive(dat, tau = tau,
                           covars = c("age", "ckd", "cirrhosis",
                                      "diabetes", "heart_failure"))
cat("Naive Cox HR (ignoring switch):", round(cox_naive$hr, 3), "\n")
cat("95% CI:", round(cox_naive$ci_low, 3), "-",
    round(cox_naive$ci_high, 3), "\n")
```

The naive Cox model uses only baseline treatment assignment and ignores switching entirely. Under non-proportional hazards, the Cox HR is a weighted average of time-varying hazard ratios that does not correspond to any single marginal risk contrast. It targets neither the treatment-policy estimand (because it is a conditional, not marginal, parameter) nor the hypothetical estimand (because post-switch outcomes are included without adjustment).

### B. Cox Model Censoring at Switch

```{r}
#| label: cox-censor-switch
cox_cs <- fit_cox_censor_switch(dat, tau = tau,
                                covars = c("age", "ckd", "cirrhosis",
                                           "diabetes", "heart_failure"))
cat("Cox HR (censor at switch):", round(cox_cs$hr, 3), "\n")
cat("95% CI:", round(cox_cs$ci_low, 3), "-",
    round(cox_cs$ci_high, 3), "\n")
```

This model censors subjects at their observed switch time. It assumes that switching is non-informative conditional on included covariates. When high-risk patients (CKD, on active treatment) preferentially switch, censoring removes the highest-risk person-time from the active arm, biasing the HR toward the null.

### C. Cox with Time-Dependent Treatment (Optional)

<!-- TODO(Joy): decide whether to include this in the main presentation -->

```{r}
#| label: cox-td
#| cache: true
cox_td <- fit_cox_td(dat, tau = tau,
                     covars = c("age", "ckd", "cirrhosis",
                                "diabetes", "heart_failure"))
cat("Cox HR (time-dependent treatment):", round(cox_td$hr, 3), "\n")
cat("95% CI:", round(cox_td$ci_low, 3), "-",
    round(cox_td$ci_high, 3), "\n")
```

Including treatment as a time-dependent covariate models the current treatment at each time point. This is sometimes used to "adjust" for switching, but it estimates a conditional HR for current treatment---not the marginal risk contrast under a target intervention. Without strong assumptions about no unmeasured time-varying confounding, this does not target the treatment-policy estimand.

### Marginal Risk Estimates from Kaplan-Meier

```{r}
#| label: km-risks
km_naive <- km_risk_at(cox_naive$survfit, t = tau)
km_cs    <- km_risk_at(cox_cs$survfit, t = tau)

cat("KM risk diff (naive, ignore switch):", round(km_naive$risk_diff, 5), "\n")
cat("KM risk diff (censor at switch):", round(km_cs$risk_diff, 5), "\n")
```

These unadjusted Kaplan-Meier estimates serve only as descriptive benchmarks and do not target any specific estimand.

::: {.callout-note}
## Why hazard ratios are fundamentally problematic as causal estimands
The problems with the Cox HR extend beyond non-collapsibility and non-proportional hazards. Even in a perfectly randomised trial, the population-level HR at time $t$ compares groups that are no longer balanced: if treatment is effective, high-risk individuals are depleted more rapidly from the control arm, so the treatment group eventually contains a larger fraction of high-risk subjects. This **built-in selection bias** means the observed HR is an attenuated version of the individual-level causal HR (Stensrud et al. 2019). Furthermore, the causal HR---conditioning on all individual-level risk factors---requires the untestable assumption that counterfactual survival times are independent conditional on covariates; without this, the HR has no causal interpretation as a hazard ratio even when the Cox model fits perfectly (Martinussen 2022). Under non-proportional hazards, the Cox HR additionally depends on the censoring distribution, making it trial-specific and difficult to compare across studies (Rufibach 2018). Survival-function contrasts such as risk differences, risk ratios, and restricted mean survival time (RMST) differences avoid these problems and provide causally interpretable, absolute effect measures.
:::

## Aligned Estimation with LMTP

The `lmtp` package implements longitudinal modified treatment policies via the sequentially doubly robust (SDR) estimator. Unlike Cox regression, `lmtp`:

- Directly targets a marginal risk under a specified intervention (not a conditional hazard ratio)
- Adjusts for time-varying confounding affected by prior treatment
- Provides doubly robust inference: consistent if either the outcome model or the treatment/censoring model is correctly specified
- Returns interpretable marginal contrasts (risk differences, risk ratios)

### Preparing Data for LMTP

```{r}
#| label: lmtp-prep
#| cache: true
#| message: false
lmtp_prep <- prepare_lmtp_data(
  dat, tau = tau, bin_width = 7,
  baseline = c("age", "sex_male", "ckd", "diabetes", "hypertension",
               "heart_failure")
)

cat("Wide data dimensions:", dim(lmtp_prep$data)[1], "x",
    dim(lmtp_prep$data)[2], "\n")
cat("Outcome columns:", length(lmtp_prep$Y_cols), "\n")
cat("Censoring columns:", length(lmtp_prep$C_cols), "\n")
```

### Running LMTP: Static Interventions

We estimate marginal risks under two static interventions:

- **Treat all with active regimen** (`static_binary_on`): set $A = 1$ for everyone
- **Treat all with comparator** (`static_binary_off`): set $A = 0$ for everyone

For the treatment-policy estimand, these interventions set baseline treatment and allow switching to occur naturally. The censoring model in LMTP adjusts for the informative censoring induced by switching.

<!-- TODO(Joy): finalise the intervention function for the hypothetical
     no-switch estimand in the longitudinal LMTP setup. -->

```{r}
#| label: lmtp-run
#| message: false
#| warning: false

# Load pre-computed LMTP results (run R/run_cached_analyses.R to generate)
lmtp_cache <- here("results", "lmtp_main.rds")
stopifnot("Run R/run_cached_analyses.R first" = file.exists(lmtp_cache))
lmtp_res <- readRDS(lmtp_cache)

cat("LMTP risk (active):", round(lmtp_res$risk_trt, 5), "\n")
cat("LMTP risk (comparator):", round(lmtp_res$risk_ctrl, 5), "\n")

rd <- lmtp_res$contrast_rd$estimates
cat("LMTP risk difference:", round(rd$estimate, 5), "\n")
cat("LMTP RD 95% CI:", round(rd$conf.low, 5), "-",
    round(rd$conf.high, 5), "\n")

rr <- lmtp_res$contrast_rr$estimates
cat("LMTP risk ratio:", round(rr$estimate, 4), "\n")
cat("LMTP RR 95% CI:", round(rr$conf.low, 5), "-",
    round(rr$conf.high, 5), "\n")
```

::: {.callout-important}
## The issue is alignment, not superiority
The Cox HR and the LMTP risk difference are not comparable as competing answers to the same question. They target different parameters on different scales. The problem is that the Cox HR is often *presented* as if it answers the estimand question, when in fact it targets a different quantity. LMTP directly targets the marginal risk under the specified intervention, which is the parameter defined by the estimand.
:::

### Practical LMTP/TMLE Specification

The analysis above uses `SL.glm` with 2-fold cross-validation for speed. For production analyses, the following guidance from the targeted learning literature (Gruber et al. 2022; Chen et al. 2023) should be followed:

- **Cross-validation folds.** The number of folds $V$ should be calibrated to the effective sample size $n_{\text{eff}}$: $V \geq 20$ when $n_{\text{eff}} < 500$; $V \geq 10$ for $500 \leq n_{\text{eff}} < 5000$; $V \geq 5$ for $5000 \leq n_{\text{eff}} < 10000$; $V \geq 2$ for $n_{\text{eff}} \geq 10000$. For binary outcomes, $n_{\text{eff}} = \min(\sum D, 5 \sum DY, 5 \sum D(1-Y))$ where $D$ indicates non-missing outcome.
- **Super Learner library.** Include at minimum a parametric model (GLM), a penalised regression (glmnet/Lasso), and one flexible learner (xgboost, BART, or GAM). A richer library reduces the risk of model misspecification.
- **Double robustness.** The SDR estimator used by `lmtp` is consistent if *either* the outcome model *or* the treatment/censoring model is correctly specified---not both. This is the key advantage over G-computation, which requires correct outcome model specification.
- **Plug-in property.** Unlike inverse probability weighting, TMLE/SDR is a substitution estimator that respects model bounds (e.g., survival probabilities remain in $[0, 1]$), avoiding the out-of-bounds estimates that can occur with IPTW in sparse data.

<!-- TODO(Joy): benchmark runtime with a richer SL library (SL.glm, SL.glmnet, SL.xgboost)
     and determine the recommended specification for the final tutorial. -->

## Known Truth from the DGP

Since we control the data-generating process, we can compute the true marginal risks under each estimand by generating large counterfactual datasets.

- **Treatment-policy truth:** generate all-treated and all-control datasets with `policy = "treatment_policy"` (switching occurs but does not censor) and no dependent censoring.
- **No-switch truth:** generate all-treated and all-control datasets with `policy = "no_switch"` (no switching at all) and no dependent censoring.

```{r}
#| label: compute-truth

# Load pre-computed ground truth (run R/run_cached_analyses.R to generate)
truth_cache <- here("results", "sim_results", "ground_truth.rds")
stopifnot("Run R/run_cached_analyses.R first" = file.exists(truth_cache))
truth_all <- readRDS(truth_cache)

cat("=== Treatment-Policy Truth ===\n")
cat("True risk (active, 180d):", round(truth_all$treatment_policy$true_risk_1, 5), "\n")
cat("True risk (comparator, 180d):", round(truth_all$treatment_policy$true_risk_0, 5), "\n")
cat("True RD:", round(truth_all$treatment_policy$true_rd, 5), "\n")
cat("True RR:", round(truth_all$treatment_policy$true_rr, 4), "\n")

cat("\n=== Hypothetical No-Switch Truth ===\n")
cat("True risk (active, 180d):", round(truth_all$no_switch$true_risk_1, 5), "\n")
cat("True risk (comparator, 180d):", round(truth_all$no_switch$true_risk_0, 5), "\n")
cat("True RD:", round(truth_all$no_switch$true_rd, 5), "\n")
cat("True RR:", round(truth_all$no_switch$true_rr, 4), "\n")
```

## Data Support Diagnostics {#sec-support}

Identification of a causal estimand may be conceptually justified, but estimation can fail in practice if the data do not provide adequate empirical support for the target intervention. This section introduces diagnostics for assessing data support and demonstrates how estimation degrades as support weakens.

**Positivity** requires that, within every stratum defined by measured confounders, there is a nonzero probability of receiving each treatment. In longitudinal settings (as with LMTP), sequential positivity requires that the cumulative product of conditional treatment and censoring probabilities remains bounded away from zero at every time point. When this product approaches zero in some strata, the corresponding inverse-probability weights become extreme and estimators become unstable.

Rather than using fixed thresholds (e.g., flagging propensity scores below 0.05 or above 0.95), a data-adaptive approach sets the positivity bound as $5 / (\sqrt{n} \cdot \log n)$, where $n$ is the sample size (Gruber et al. 2022). For $n = 2000$, this gives bounds of approximately $(0.025, 0.975)$; for $n = 500$, approximately $(0.036, 0.964)$. The `compute_ps_diagnostics()` function in our support diagnostics reports both conventional and adaptive thresholds.

### Three Support Scenarios

We create three versions of the DGP by modifying parameters that control switching intensity and confounding.

```{r}
#| label: support-scenarios
#| cache: true

set.seed(101)
dat_good <- generate_hep_data(
  N = 2000, np_hazard = TRUE, dep_censor = TRUE,
  complexity = TRUE, policy = "treatment_policy", seed = 101
)

set.seed(102)
dat_strained <- generate_hep_data(
  N = 2000, np_hazard = TRUE, dep_censor = TRUE,
  complexity = TRUE, policy = "treatment_policy",
  gamma_A = 1.5, gamma_ckd = 1.2, lambda_sw0 = 5e-5,
  seed = 102
)

set.seed(103)
dat_poor <- generate_hep_data(
  N = 2000, np_hazard = TRUE, dep_censor = TRUE,
  complexity = TRUE, policy = "treatment_policy",
  gamma_A = 2.5, gamma_ckd = 2.0, lambda_sw0 = 1e-4,
  seed = 103
)
```

### Propensity Score Overlap

```{r}
#| label: fig-support-ps
#| fig-cap: "Data support diagnostic: propensity score overlap across three support scenarios. As confounding strengthens, overlap deteriorates."
#| fig-height: 9
#| fig-width: 7

scenario_list <- list(
  "1. Good support" = dat_good,
  "2. Strained support" = dat_strained,
  "3. Poor support" = dat_poor
)

plot_support_scenarios(scenario_list)
```

### Support Summary Table

```{r}
#| label: tbl-support
#| tbl-cap: "Support diagnostic summary across scenarios"

support_tbl <- bind_rows(
  summarize_support(dat_good, "Good support"),
  summarize_support(dat_strained, "Strained support"),
  summarize_support(dat_poor, "Poor support")
)

kable(support_tbl, digits = 3, format = "pipe")
```

### How Estimation Degrades with Weak Support

```{r}
#| label: support-estimation
#| message: false
#| warning: false

# Load pre-computed support results (run R/run_cached_analyses.R to generate)
support_cache <- here("results", "support_estimation.rds")
stopifnot("Run R/run_cached_analyses.R first" = file.exists(support_cache))
support_results <- readRDS(support_cache)

kable(support_results, digits = 4, format = "pipe")
```

Under good support, all methods produce estimates. As support degrades:

- Cox HRs become unstable and may shift toward the null (censor-at-switch removes the highest-risk patients)
- LMTP estimates may show larger standard errors, instability, or convergence warnings
- The practical lesson: identification may be conceptually valid while estimation fails due to lack of empirical support

::: {.callout-warning}
## Identification vs. estimation
A target estimand can be well-defined and conceptually identified under the assumed causal model, but estimation can still fail if the data do not contain enough subjects who follow (or could plausibly follow) the target intervention. Always check support diagnostics before interpreting estimates.
:::

## Repeated Simulation Study

We now assess estimator performance over repeated samples from the DGP. This reveals bias, variance, and coverage properties that a single dataset cannot show. For each estimand, we compare misaligned Cox estimators to the aligned LMTP estimator.

<!-- TODO(Joy): fill in exact runtime notes after benchmarking -->
<!-- TODO(Joy): benchmark whether 200 iterations is feasible or reduce for development -->

```{r}
#| label: run-sim
#| message: false
#| warning: false

source(here("R", "run_simulations.R"))

# Load pre-computed simulation results (run R/run_cached_analyses.R to generate)
sim_cache <- here("results", "sim_study_main.rds")
stopifnot("Run R/run_cached_analyses.R first" = file.exists(sim_cache))
sim_results <- readRDS(sim_cache)
```

### Results by Estimand

```{r}
#| label: tbl-sim-tp
#| tbl-cap: "Simulation results: Treatment-policy estimand (50 iterations, N = 1000)"

sim_tp <- sim_results %>% filter(estimand == "treatment_policy")
sim_summary_tp <- summarize_simulation(
  sim_tp,
  truth_rd = truth_all$treatment_policy$true_rd
)

kable(sim_summary_tp %>% select(-estimand), digits = 4, format = "pipe")
```

```{r}
#| label: tbl-sim-ns
#| tbl-cap: "Simulation results: Hypothetical no-switch estimand (50 iterations, N = 1000)"

sim_ns <- sim_results %>% filter(estimand == "no_switch")
sim_summary_ns <- summarize_simulation(
  sim_ns,
  truth_rd = truth_all$no_switch$true_rd
)

kable(sim_summary_ns %>% select(-estimand), digits = 4, format = "pipe")
```

### Performance Plot

```{r}
#| label: fig-sim-performance
#| fig-cap: "Distribution of estimates across simulation iterations, by estimand and method. Red dashed line = true risk difference."
#| fig-height: 6
#| fig-width: 9

sim_plot_data <- sim_results %>%
  filter(converged) %>%
  mutate(
    estimate = case_when(
      method == "LMTP SDR" ~ risk_diff,
      TRUE ~ hr
    ),
    scale = case_when(
      method == "LMTP SDR" ~ "Risk Difference",
      TRUE ~ "Hazard Ratio"
    ),
    estimand_label = case_when(
      estimand == "treatment_policy" ~ "Treatment-Policy",
      estimand == "no_switch" ~ "Hypothetical No-Switch"
    )
  )

truth_lines <- tibble(
  estimand_label = c("Treatment-Policy", "Hypothetical No-Switch"),
  scale = "Risk Difference",
  xint = c(truth_all$treatment_policy$true_rd,
           truth_all$no_switch$true_rd)
)

ggplot(sim_plot_data, aes(x = estimate, fill = method)) +
  geom_histogram(alpha = 0.6, bins = 25, position = "identity") +
  facet_grid(estimand_label ~ scale, scales = "free") +
  geom_vline(data = truth_lines,
             aes(xintercept = xint), linetype = "dashed", color = "red") +
  labs(x = "Estimate", y = "Count", fill = "Method",
       title = "Estimator Performance Within Each Estimand") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

```{r}
#| label: sim-convergence
conv_summary <- sim_results %>%
  group_by(estimand, method) %>%
  summarise(
    total = n(),
    converged = sum(converged),
    failed = sum(!converged),
    .groups = "drop"
  )
cat("Convergence summary:\n")
print(conv_summary)
```

## Interpretation

The results demonstrate a consistent pattern:

1. **The naive Cox model** (ignoring switching) produces a hazard ratio that reflects a mixture of the treatment effect and the effect of switching-induced changes to the risk set. Under non-proportional hazards, this HR does not correspond to any single marginal risk contrast. It is not aligned with either estimand.

2. **The Cox model censoring at switch** is biased under informative switching. By removing the highest-risk patients from the treatment arm, it attenuates the apparent treatment effect. It does not correctly target the hypothetical no-switch estimand because the censoring-at-switch assumption (non-informative censoring conditional on covariates) is violated.

3. **LMTP** directly targets the marginal risk under each static intervention and adjusts for confounding and censoring through its doubly robust longitudinal estimation. When data support is adequate, LMTP recovers the truth with correct coverage.

4. **Data support matters.** Even when the estimand is well-defined, estimation degrades when empirical support is inadequate. Propensity score overlap and switching rate diagnostics should be examined before trusting any estimate.

The central lesson is not that LMTP is generically superior to Cox regression. It is that **LMTP is better aligned with the estimand** when the estimand is defined as a marginal risk contrast under a longitudinal intervention, and when time-varying confounding affected by prior treatment is present.

## Sensitivity to Unmeasured Confounding

Even with a well-aligned estimator and adequate data support, the causal interpretation of any observational estimate rests on the assumption of no unmeasured confounding (exchangeability). The **causal gap** $\delta = \psi_{\text{stat}} - \psi_{\text{causal}}$ quantifies the difference between what the statistical estimand identifies and the true causal quantity. In a randomised trial, $\delta = 0$ by design; in an observational study, $\delta$ reflects residual confounding from unmeasured common causes of treatment and outcome.

A useful diagnostic is the **G-value** (Gruber et al. 2024): the minimum causal gap needed to shift the confidence interval to include the null. If the G-value is large relative to the magnitude of the measured confounding adjustment (the difference between unadjusted and adjusted estimates), the finding is robust. If it is small, even modest unmeasured confounding could nullify the result. Sensitivity analyses should examine how the point estimate and confidence interval change across a range of hypothesised $\delta$ values, typically from 0 to 5 times the measured confounding adjustment (Dang et al. 2023).

This tutorial does not implement a full sensitivity analysis, but the causal roadmap framework requires it as a distinct step (Step 6). In a complete analysis, a sensitivity plot showing shifted estimates and confidence bounds across a range of causal gap values should accompany every primary result.

<!-- TODO(Joy): implement a nonparametric sensitivity analysis function and
     add a worked example to the simulation study. -->

## Key Messages

1. **The main problem is estimand-estimator misalignment.** Analysts often run regression procedures that do not target the stated estimand when intercurrent events are present.

2. **Cox models that censor at switching** implicitly assume non-informative censoring conditional on included covariates. When switching is driven by evolving covariates affected by prior treatment, this assumption is generally not plausible.

3. **Hazard ratios are not risk contrasts.** Under non-proportional hazards, competing risks, or time-varying treatment effects, the Cox HR does not correspond to a marginal risk difference or ratio. Estimands defined as risk contrasts require estimators that target risk contrasts.

4. **LMTP aligns estimation with the longitudinal intervention** defining the estimand. It adjusts for time-varying confounding under the identifying assumptions and returns interpretable marginal contrasts.

5. **Support diagnostics are mandatory.** An estimand can be conceptually identified while estimation fails due to lack of data support. Always check propensity score overlap, switching rates by stratum, and effective sample sizes before interpreting estimates.

6. **Report transparently.** State the estimand, the assumptions required for identification, the diagnostics that support (or fail to support) those assumptions, and the alignment between the estimator and the estimand.

## Sources

- ICH E9(R1) Addendum (2019). Addendum on estimands and sensitivity analyses in clinical trials.
- Williams N, Diaz I (2023). lmtp: Non-parametric causal effects of feasible interventions based on modified treatment policies. R package.
- van der Laan MJ, Rose S (2011). Targeted Learning: Causal Inference for Observational and Experimental Data.
- Hernan MA, Robins JM (2020). Causal Inference: What If.
- Dang LE, Tarp JM, Abrahamsen B, et al. (2023). A causal roadmap for generating high-quality real-world evidence. Journal of Clinical and Translational Science.
- Gruber S, Lee H, Phillips R, Ho M, van der Laan MJ (2022). Developing a targeted learning-based statistical analysis plan. Statistics in Biopharmaceutical Research.
- Gruber S, Phillips RV, Lee H, van der Laan MJ (2024). Targeted learning: toward a future informed by real-world evidence. Statistics in Biopharmaceutical Research.
- Chen RC, Rosenblum M, van der Laan MJ, Gruber S (2023). Beyond the Cox hazard ratio: a targeted learning approach to survival analysis in a cardiovascular outcome trial application. Clinical Trials.
- Martinussen T (2022). Causality and the Cox regression model. Annual Review of Statistics and Its Application.
- Stensrud MJ, Aalen JM, Aalen OO, Valberg M (2019). Limitations of hazard ratios in clinical trials. European Heart Journal.
- Rufibach K (2019). Treatment effect quantification for time-to-event endpoints: estimands, analysis strategies, and beyond. Pharmaceutical Statistics.
- Petersen ML (2014). Applying a causal road map in settings with time-dependent confounding. Epidemiology.
