---
title: "Estimand-Estimator Alignment in Time-to-Event Safety Analyses"
subtitle: "A hepatitis B case study with longitudinal targeted learning"
format: html
---

<!--
TODO(Joy): verify exact variable mapping from DGP.R to chapter notation.
TODO(Joy): refine hepatitis B DGP parameters to match realistic renal-failure
  incidence in tenofovir vs entecavir from the literature.
TODO(Joy): replace placeholder intervention function with the final
  study-specific rule if needed.
TODO(Joy): decide whether to include the time-dependent Cox model in the
  main presentation or relegate it to an appendix.
-->

# Estimand-Estimator Alignment in Time-to-Event Safety Analyses

## Introduction

In real-world safety studies, the question "Does this drug increase the risk of adverse events?" requires a precise definition of the causal quantity being estimated. The ICH E9(R1) addendum formalises this through the **estimand framework**, which requires specifying how intercurrent events---treatment switching, discontinuation, death from other causes---are handled. Each strategy defines a different estimand, and each estimand demands a compatible estimator.

The central problem this chapter addresses is not which estimand to choose. Rather, it is that **conventional regression analyses often fail to target the stated estimand** when intercurrent events are present. A naive Cox model that ignores switching, or one that censors at switching, may appear to estimate a treatment effect, but the quantity it targets can be misaligned with the scientific question. Longitudinal targeted learning via `lmtp` offers a principled alternative that directly targets the intervention-defined estimand and adjusts for time-varying confounding.

This chapter focuses on two estimands---the **treatment-policy** estimand and the **hypothetical no-switch** estimand---and demonstrates how to align (or misalign) estimators with each. The goal is not to compare estimands against each other, but to ensure that the chosen estimator matches the causal question.

This chapter follows the structure of the **causal roadmap** for generating high-quality real-world evidence (Dang et al. 2023; Petersen 2014). The roadmap proceeds through: (1) defining the causal question and estimand as a function of counterfactual outcomes---not as a regression coefficient; (2) specifying a causal model (DAG); (3) assessing identifiability under exchangeability, positivity, and consistency; (4) defining the statistical estimand; (5) choosing an estimator; and (6) conducting sensitivity analyses. This tutorial concentrates on steps 1, 3--5: defining estimands, diagnosing support, and aligning estimators. A key principle of the roadmap is that the estimand should be defined mathematically before any statistical method is chosen, and that the gap between the causal estimand and the statistical estimand should be made explicit.

**Goals:**

1. Define the treatment-policy and hypothetical no-switch estimands under the ICH E9(R1) framework
2. Show how conventional Cox regression analyses can fail to target these estimands
3. Demonstrate how `lmtp`-based longitudinal targeted learning can better align estimation with the stated estimand
4. Diagnose when empirical support for a target intervention is inadequate

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  fig.align = "center",
  cache = TRUE
)
```

```{r}
#| label: load-packages
#| message: false
#| warning: false
library(tidyverse)
library(survival)
library(here)
library(knitr)
```

```{r}
#| label: source-helpers
#| message: false
#| warning: false
source(here("DGP.R"))
source(here("R", "helpers.R"))
source(here("R", "support_diagnostics.R"))

# Check that cached results exist (run R/run_cached_analyses.R first)
cache_files <- c(
  here("results", "lmtp_main.rds"),
  here("results", "support_estimation.rds"),
  here("results", "sim_study_main.rds"),
  here("results", "sim_results", "ground_truth.rds")
)
missing <- cache_files[!file.exists(cache_files)]
if (length(missing) > 0) {
  warning("Missing cached results. Run R/run_cached_analyses.R first:\n",
          paste(" ", basename(missing), collapse = "\n"))
}
```

## The Hepatitis B Safety Example

We use a simulated post-marketing safety study comparing two hepatitis B antiviral regimens for the risk of **renal failure** (the safety endpoint). The setting is motivated by real concerns about nephrotoxicity associated with tenofovir-containing regimens relative to entecavir-based alternatives in patients with chronic hepatitis B.

Key features of the example:

- **Two treatments:** an active regimen (analogous to tenofovir-containing therapy) and a comparator (analogous to entecavir-based therapy), with confounded treatment assignment driven by baseline renal function (CKD), cirrhosis severity, and age.
- **Intercurrent event:** informative treatment switching. Patients on the active regimen who develop early signs of renal impairment may switch to the comparator. Because CKD drives both switching and renal failure, switching is informative.
- **Outcome:** time to renal failure (or a composite renal endpoint), observed over a 180-day risk window.
- **Baseline confounders:** demographics (age, sex, race), comorbidities (CKD, diabetes, hypertension, cirrhosis, heart failure), and co-medications (NSAIDs, ACE/ARBs, statins).

This example illustrates a realistic scenario in which informative switching invalidates naive regression approaches and where the choice of estimand strategy has concrete consequences for the analysis.

<!-- TODO(Joy): add one paragraph with real-data references if a project-approved
     example from the hepatitis B nephrotoxicity literature is available. -->

## The Regulatory Context: ICH E9(R1)

The ICH E9(R1) addendum (2019) requires explicit specification of five elements for any clinical or post-marketing estimand:

1. **The population** of interest
2. **The treatment** conditions being compared
3. **The outcome** variable
4. **The intercurrent events** that may occur after treatment initiation
5. **The strategy** for handling each intercurrent event

When 15--20% of patients on the active regimen switch due to early renal signals, and when patients with CKD are more likely to both switch and develop renal failure, the strategy used to handle switching defines fundamentally different estimands---and different estimands require different estimators.

## The Data-Generating Process

We use `generate_hep_data()` from the project DGP to simulate realistic post-marketing safety data. The DGP produces a cohort of patients eligible for one of two hepatitis B treatments, with renal failure as the safety endpoint.

| DGP Component | Description |
|---|---|
| **Baseline covariates** | Demographics (age, sex, race); comorbidities (CKD, diabetes, hypertension, cirrhosis, heart failure); co-medications (NSAIDs, ACE/ARBs, statins) |
| **Treatment assignment** | Active vs comparator, confounded by CKD, cirrhosis, age, and (when `complexity=TRUE`) nonlinear covariate interactions |
| **Event process** | Piecewise exponential renal-failure hazard; when `np_hazard=TRUE`, HR = 1.25 early and HR = 0.70 late (non-proportional hazards) |
| **Switching process** | Hazard-based informative switching driven by treatment and CKD |
| **Censoring process** | When `dep_censor=TRUE`, censoring depends on baseline risk; otherwise administrative censoring only |

The DGP accepts a `policy` argument:

- `policy = "treatment_policy"`: switching occurs but does **not** censor. Subjects are followed until event or administrative censoring regardless of switching.
- `policy = "no_switch"`: no switching occurs. This generates potential outcomes under the hypothetical world where nobody switches.

<!-- TODO(Joy): verify exact variable mapping from DGP.R to chapter notation -->

```{r}
#| label: generate-data
#| cache: true
set.seed(2026)
dat <- generate_hep_data(
  N           = 10000,
  np_hazard   = TRUE,
  dep_censor  = TRUE,
  complexity  = TRUE,
  policy      = "treatment_policy",
  seed        = 2026
)

cat("N:", nrow(dat), "\n")
cat("Event rate:", round(mean(dat$event), 4), "\n")
cat("Switch rate:", round(mean(dat$switched), 4), "\n")
cat("Switch rate (active + CKD):",
    round(mean(dat$switched[dat$treatment == 1 & dat$ckd == 1]), 4), "\n")
cat("Switch rate (comparator):",
    round(mean(dat$switched[dat$treatment == 0]), 4), "\n")
cat("Median follow-up:", round(median(dat$follow_time), 1), "days\n")
```

Switching is driven by both treatment assignment and CKD status. Patients on the active regimen with CKD switch at the highest rate. Because CKD also increases renal failure risk, switching is **informative**: the patients who switch are those at greatest risk for the outcome.

```{r}
#| label: fig-dag
#| fig-cap: "Causal structure: baseline confounders (W) affect treatment, switching, censoring, and renal failure. Switching is informative because W (especially CKD) drives both switching and the outcome."
#| fig-width: 8
#| fig-height: 5
library(dagitty)
library(ggdag)

dag_safety <- dagify(
  Treatment ~ W,
  Switch ~ Treatment + W,
  Censoring ~ Switch + W,
  RenalFailure ~ Treatment + Switch + W,
  exposure = "Treatment", outcome = "RenalFailure",
  labels = c(W = "Baseline\nconfounders\n(CKD, age, ...)",
             Treatment = "Treatment\n(active vs comparator)",
             Switch = "Regimen\nswitch",
             Censoring = "Censoring",
             RenalFailure = "Renal\nfailure"),
  coords = list(
    x = c(W = 1, Treatment = 0, Switch = 1, Censoring = 2, RenalFailure = 1.5),
    y = c(W = 1, Treatment = 0, Switch = 0, Censoring = 0, RenalFailure = -1)
  )
)
ggdag(dag_safety, text = FALSE, use_labels = "label", text_size = 3) +
  theme_dag()
```

## Five Estimand Strategies Under ICH E9(R1)

The ICH E9(R1) addendum defines five strategies for handling intercurrent events. In a safety study where the primary intercurrent event is treatment switching, each strategy answers a different clinical question. We define and simulate all five below.

### Estimand 1: Treatment-Policy

**Question:** What is the risk of renal failure by 180 days comparing populations *initiated* on the active regimen versus the comparator, regardless of subsequent treatment changes?

**Intercurrent event strategy:** Ignore switching. Follow all patients from initiation regardless of what happens to their treatment.

**Target parameter:** $\psi^{TP} = E[Y^{a=1}(180)] - E[Y^{a=0}(180)]$, where $Y^a(t)$ is the counterfactual renal-failure indicator under baseline initiation of regime $a$, with all subsequent intercurrent events occurring as they naturally would.

### Estimand 2: Hypothetical No-Switch

**Question:** What *would* the risk of renal failure be by 180 days if, contrary to fact, no patients switched regimens?

**Intercurrent event strategy:** Model the hypothetical scenario in which switching does not occur. This requires identifying and adjusting for the time-varying confounders that drive both switching and the outcome.

**Target parameter:** $\psi^{HYP} = E[Y^{a=1, \text{no switch}}(180)] - E[Y^{a=0, \text{no switch}}(180)]$

### Estimand 3: While-on-Treatment

**Question:** What is the risk of renal failure while patients remain on their assigned treatment?

**Intercurrent event strategy:** Censor at the time of switching. Only events occurring before any switch contribute to the analysis.

**Target parameter:** $\psi^{WOT} = E[Y^{a=1}(180) \mid \text{no switch by } 180] - E[Y^{a=0}(180) \mid \text{no switch by } 180]$

This estimand is clinically relevant: it captures the treatment effect during the period patients can actually tolerate therapy. When switching is informative, naive estimation (simply censoring at switch in a Cox model) is particularly susceptible to bias because the patients removed are those at highest risk. Inverse probability of censoring weighting (IPCW) or LMTP with a censoring model for switching can reduce this bias.

### Estimand 4: Composite

**Question:** What is the risk of renal failure *or* treatment switching (whichever comes first) by 180 days?

**Intercurrent event strategy:** Absorb switching into the outcome. Switching is treated as a component of a composite endpoint alongside renal failure.

**Target parameter:** $\psi^{COMP} = E[Y_{comp}^{a=1}(180)] - E[Y_{comp}^{a=0}(180)]$, where $Y_{comp}^a(t) = 1$ if renal failure or switching occurs by time $t$ under regime $a$.

This strategy avoids the need to model switching, since it is no longer an intercurrent event but part of the outcome. The trade-off is interpretability: the composite conflates treatment tolerability (switching due to toxicity) with the safety outcome (renal failure).

### Estimand 5: Principal Stratum

**Question:** What is the risk of renal failure by 180 days among patients who *would not switch* under either treatment assignment?

**Intercurrent event strategy:** Restrict to the principal stratum of "never-switchers"---patients who would remain on treatment regardless of which drug they received.

**Target parameter:** $\psi^{PS} = E[Y^{a=1}(180) \mid S^{a=1} = 0, S^{a=0} = 0] - E[Y^{a=0}(180) \mid S^{a=1} = 0, S^{a=0} = 0]$, where $S^a$ indicates switching under regime $a$.

The principal stratum is defined by joint potential outcomes for switching and is not directly observable. Identification requires strong assumptions (e.g., monotonicity: if a patient would not switch on the active drug, they would not switch on the comparator). In the simulation, we can identify the true never-switcher stratum because we control the DGP.

### Summary

| Strategy | Handling | Key assumption | Primary trade-off |
|---|---|---|---|
| **Treatment-policy** | Ignore switching | None for switching | Includes post-switch outcomes |
| **Hypothetical** | Model world without switching | Correct switching model | Requires modelling the switching mechanism |
| **While-on-treatment** | Censor at switch | Non-informative censoring (or IPCW) | Bias if switching is informative |
| **Composite** | Include switching in outcome | None for switching | Conflates switching and renal failure |
| **Principal stratum** | Restrict to never-switchers | Monotonicity | Targets a latent subpopulation |

## Estimand-Estimator Alignment

The table below is the conceptual core of this chapter. For each estimand, it lists the aligned LMTP estimator and the misaligned Cox estimators, together with the reason for misalignment.

```{r}
#| label: tbl-alignment
#| tbl-cap: "Estimand versus estimator: aligned and misaligned approaches"

alignment_tbl <- tibble::tribble(
  ~Estimand, ~Estimator, ~Aligned, ~`Reason for misalignment`,
  "Treatment-policy", "LMTP SDR (static on/off)", "Yes",
    "Targets marginal risk under baseline initiation; adjusts for censoring",
  "Treatment-policy", "Cox naive (ignore switch)", "No",
    "Targets conditional HR, not marginal RD; risk set altered by informative switching",
  "Treatment-policy", "Cox censor-at-switch", "No",
    "Censoring at switch removes high-risk subjects; wrong estimand entirely",
  "Treatment-policy", "Cox time-dependent trt", "No",
    "Conditional HR with time-varying exposure; does not target marginal risk",
  "Hypothetical no-switch", "LMTP SDR (no-switch intervention)", "Yes",
    "Targets marginal risk in a world without switching; adjusts for confounding",
  "Hypothetical no-switch", "Cox naive (ignore switch)", "No",
    "Includes post-switch outcomes; does not model the hypothetical world",
  "Hypothetical no-switch", "Cox censor-at-switch", "No",
    "Assumes non-informative censoring; biased when switching is informative",
  "While-on-treatment", "LMTP SDR (censor model for switch)", "Yes",
    "Targets marginal risk while on treatment; upweights patients who stay despite high switch propensity",
  "While-on-treatment", "Cox censor-at-switch", "No",
    "Assumes non-informative censoring; biased when high-risk patients preferentially switch",
  "While-on-treatment", "Cox naive (ignore switch)", "No",
    "Includes post-switch person-time; does not restrict to on-treatment experience",
  "Composite", "LMTP SDR (composite outcome)", "Yes",
    "Targets marginal risk of composite (renal failure or switch); no switching model needed",
  "Composite", "Cox naive (composite outcome)", "Partial",
    "HR for composite endpoint; conditional not marginal, but switching is not an ICE",
  "Principal stratum", "LMTP SDR (restricted to never-switchers)", "Approximate",
    "Requires identifying the latent never-switcher stratum; strong assumptions",
  "Principal stratum", "Cox naive (subset analysis)", "No",
    "Observed non-switchers are not the same as the principal stratum"
)

kable(alignment_tbl, format = "pipe")
```

A brief conceptual explanation: when switching is informative (driven by CKD and treatment), censoring at switch removes the highest-risk person-time from the active treatment arm. This makes the active treatment appear safer than it is. The naive Cox model, by contrast, includes post-switch outcomes that may reflect the comparator treatment rather than the active one. Neither targets the marginal risk contrast that defines the estimand. LMTP avoids these problems by directly targeting the counterfactual risk under a specified intervention and adjusting for confounding and censoring through its doubly robust longitudinal estimation.

## Conventional Regression Analyses (Misaligned)

### A. Naive Baseline Cox Model

```{r}
#| label: cox-naive
tau <- 180

cox_naive <- fit_cox_naive(dat, tau = tau,
                           covars = c("age", "ckd", "cirrhosis",
                                      "diabetes", "heart_failure"))
cat("Naive Cox HR (ignoring switch):", round(cox_naive$hr, 3), "\n")
cat("95% CI:", round(cox_naive$ci_low, 3), "-",
    round(cox_naive$ci_high, 3), "\n")
```

The naive Cox model uses only baseline treatment assignment and ignores switching entirely. Under non-proportional hazards, the Cox HR is a weighted average of time-varying hazard ratios that does not correspond to any single marginal risk contrast. It targets neither the treatment-policy estimand (because it is a conditional, not marginal, parameter) nor the hypothetical estimand (because post-switch outcomes are included without adjustment).

### B. Cox Model Censoring at Switch

```{r}
#| label: cox-censor-switch
cox_cs <- fit_cox_censor_switch(dat, tau = tau,
                                covars = c("age", "ckd", "cirrhosis",
                                           "diabetes", "heart_failure"))
cat("Cox HR (censor at switch):", round(cox_cs$hr, 3), "\n")
cat("95% CI:", round(cox_cs$ci_low, 3), "-",
    round(cox_cs$ci_high, 3), "\n")
```

This model censors subjects at their observed switch time. It assumes that switching is non-informative conditional on included covariates. When high-risk patients (CKD, on active treatment) preferentially switch, censoring removes the highest-risk person-time from the active arm, biasing the HR toward the null.

### C. Cox with Time-Dependent Treatment (Optional)

<!-- TODO(Joy): decide whether to include this in the main presentation -->

```{r}
#| label: cox-td
#| cache: true
cox_td <- fit_cox_td(dat, tau = tau,
                     covars = c("age", "ckd", "cirrhosis",
                                "diabetes", "heart_failure"))
cat("Cox HR (time-dependent treatment):", round(cox_td$hr, 3), "\n")
cat("95% CI:", round(cox_td$ci_low, 3), "-",
    round(cox_td$ci_high, 3), "\n")
```

Including treatment as a time-dependent covariate models the current treatment at each time point. This is sometimes used to "adjust" for switching, but it estimates a conditional HR for current treatment---not the marginal risk contrast under a target intervention. Without strong assumptions about no unmeasured time-varying confounding, this does not target the treatment-policy estimand.

### Marginal Risk Estimates from Kaplan-Meier

```{r}
#| label: km-risks
km_naive <- km_risk_at(cox_naive$survfit, t = tau)
km_cs    <- km_risk_at(cox_cs$survfit, t = tau)

cat("KM risk diff (naive, ignore switch):", round(km_naive$risk_diff, 5), "\n")
cat("KM risk diff (censor at switch):", round(km_cs$risk_diff, 5), "\n")
```

These unadjusted Kaplan-Meier estimates serve only as descriptive benchmarks and do not target any specific estimand.

::: {.callout-note}
## Why hazard ratios are fundamentally problematic as causal estimands
The problems with the Cox HR extend beyond non-collapsibility and non-proportional hazards. Even in a perfectly randomised trial, the population-level HR at time $t$ compares groups that are no longer balanced: if treatment is effective, high-risk individuals are depleted more rapidly from the control arm, so the treatment group eventually contains a larger fraction of high-risk subjects. This **built-in selection bias** means the observed HR is an attenuated version of the individual-level causal HR (Stensrud et al. 2019). Furthermore, the causal HR---conditioning on all individual-level risk factors---requires the untestable assumption that counterfactual survival times are independent conditional on covariates; without this, the HR has no causal interpretation as a hazard ratio even when the Cox model fits perfectly (Martinussen 2022). Under non-proportional hazards, the Cox HR additionally depends on the censoring distribution, making it trial-specific and difficult to compare across studies (Rufibach 2018). Survival-function contrasts such as risk differences, risk ratios, and restricted mean survival time (RMST) differences avoid these problems and provide causally interpretable, absolute effect measures.
:::

## Aligned Estimation with LMTP

The `lmtp` package implements longitudinal modified treatment policies via the sequentially doubly robust (SDR) estimator. Unlike Cox regression, `lmtp`:

- Directly targets a marginal risk under a specified intervention (not a conditional hazard ratio)
- Adjusts for time-varying confounding affected by prior treatment
- Provides doubly robust inference: consistent if either the outcome model or the treatment/censoring model is correctly specified
- Returns interpretable marginal contrasts (risk differences, risk ratios)

### Preparing Data for LMTP

```{r}
#| label: lmtp-prep
#| cache: true
#| message: false
lmtp_prep <- prepare_lmtp_data(
  dat, tau = tau, bin_width = 14,
  baseline = c("age", "sex_male", "ckd", "diabetes", "hypertension",
               "heart_failure")
)

cat("Wide data dimensions:", dim(lmtp_prep$data)[1], "x",
    dim(lmtp_prep$data)[2], "\n")
cat("Outcome columns:", length(lmtp_prep$Y_cols), "\n")
cat("Censoring columns:", length(lmtp_prep$C_cols), "\n")
```

### Running LMTP: Static Interventions

We estimate marginal risks under two static interventions:

- **Treat all with active regimen** (`static_binary_on`): set $A = 1$ for everyone
- **Treat all with comparator** (`static_binary_off`): set $A = 0$ for everyone

For the treatment-policy estimand, these interventions set baseline treatment and allow switching to occur naturally. The censoring model in LMTP adjusts for the informative censoring induced by switching.

<!-- TODO(Joy): finalise the intervention function for the hypothetical
     no-switch estimand in the longitudinal LMTP setup. -->

```{r}
#| label: lmtp-run
#| message: false
#| warning: false

# Load pre-computed LMTP results (run R/run_cached_analyses.R to generate)
lmtp_cache <- here("results", "lmtp_main.rds")
stopifnot("Run R/run_cached_analyses.R first" = file.exists(lmtp_cache))
lmtp_res <- readRDS(lmtp_cache)

cat("LMTP risk (active):", round(lmtp_res$risk_trt, 5), "\n")
cat("LMTP risk (comparator):", round(lmtp_res$risk_ctrl, 5), "\n")

rd <- lmtp_res$contrast_rd$estimates
cat("LMTP risk difference:", round(rd$estimate, 5), "\n")
cat("LMTP RD 95% CI:", round(rd$conf.low, 5), "-",
    round(rd$conf.high, 5), "\n")

rr <- lmtp_res$contrast_rr$estimates
cat("LMTP risk ratio:", round(rr$estimate, 4), "\n")
cat("LMTP RR 95% CI:", round(rr$conf.low, 5), "-",
    round(rr$conf.high, 5), "\n")
```

::: {.callout-important}
## The issue is alignment, not superiority
The Cox HR and the LMTP risk difference are not comparable as competing answers to the same question. They target different parameters on different scales. The problem is that the Cox HR is often *presented* as if it answers the estimand question, when in fact it targets a different quantity. LMTP directly targets the marginal risk under the specified intervention, which is the parameter defined by the estimand.
:::

### Practical LMTP/TMLE Specification

The analysis above uses `SL.glm` with 2-fold cross-validation for speed. For production analyses, the following guidance from the targeted learning literature (Gruber et al. 2022; Chen et al. 2023) should be followed:

- **Cross-validation folds.** The number of folds $V$ should be calibrated to the effective sample size $n_{\text{eff}}$: $V \geq 20$ when $n_{\text{eff}} < 500$; $V \geq 10$ for $500 \leq n_{\text{eff}} < 5000$; $V \geq 5$ for $5000 \leq n_{\text{eff}} < 10000$; $V \geq 2$ for $n_{\text{eff}} \geq 10000$. For binary outcomes, $n_{\text{eff}} = \min(\sum D, 5 \sum DY, 5 \sum D(1-Y))$ where $D$ indicates non-missing outcome.
- **Super Learner library.** Include at minimum a parametric model (GLM), a penalised regression (glmnet/Lasso), and one flexible learner (xgboost, BART, or GAM). A richer library reduces the risk of model misspecification.
- **Double robustness.** The SDR estimator used by `lmtp` is consistent if *either* the outcome model *or* the treatment/censoring model is correctly specified---not both. This is the key advantage over G-computation, which requires correct outcome model specification.
- **Plug-in property.** Unlike inverse probability weighting, TMLE/SDR is a substitution estimator that respects model bounds (e.g., survival probabilities remain in $[0, 1]$), avoiding the out-of-bounds estimates that can occur with IPTW in sparse data.

<!-- TODO(Joy): benchmark runtime with a richer SL library (SL.glm, SL.glmnet, SL.xgboost)
     and determine the recommended specification for the final tutorial. -->

## Known Truth from the DGP

Since we control the data-generating process, we can compute the true marginal risks under each estimand by generating large counterfactual datasets.

- **Treatment-policy truth:** generate all-treated and all-control datasets with `policy = "treatment_policy"` (switching occurs but does not censor) and no dependent censoring.
- **No-switch truth:** generate all-treated and all-control datasets with `policy = "no_switch"` (no switching at all) and no dependent censoring.

```{r}
#| label: compute-truth

# Load pre-computed ground truth (run R/run_cached_analyses.R to generate)
truth_cache <- here("results", "sim_results", "ground_truth.rds")
stopifnot("Run R/run_cached_analyses.R first" = file.exists(truth_cache))
truth_all <- readRDS(truth_cache)

truth_summary <- tibble::tibble(
  Estimand = c("Treatment-policy", "Hypothetical no-switch",
               "While-on-treatment", "Composite", "Principal stratum"),
  `Risk (active)` = c(truth_all$treatment_policy$true_risk_1,
                       truth_all$no_switch$true_risk_1,
                       truth_all$while_on_treatment$true_risk_1,
                       truth_all$composite$true_risk_1,
                       truth_all$principal_stratum$true_risk_1),
  `Risk (comparator)` = c(truth_all$treatment_policy$true_risk_0,
                           truth_all$no_switch$true_risk_0,
                           truth_all$while_on_treatment$true_risk_0,
                           truth_all$composite$true_risk_0,
                           truth_all$principal_stratum$true_risk_0),
  `Risk Difference` = c(truth_all$treatment_policy$true_rd,
                         truth_all$no_switch$true_rd,
                         truth_all$while_on_treatment$true_rd,
                         truth_all$composite$true_rd,
                         truth_all$principal_stratum$true_rd),
  `Risk Ratio` = c(truth_all$treatment_policy$true_rr,
                    truth_all$no_switch$true_rr,
                    truth_all$while_on_treatment$true_rr,
                    truth_all$composite$true_rr,
                    truth_all$principal_stratum$true_rr),
  `True HR` = c(truth_all$treatment_policy$true_hr,
                truth_all$no_switch$true_hr,
                truth_all$while_on_treatment$true_hr,
                truth_all$composite$true_hr,
                truth_all$principal_stratum$true_hr)
)

kable(truth_summary, digits = 5, format = "pipe",
      caption = "True 180-day risks and hazard ratios by estimand (Monte Carlo, N = 500,000)")
```

### What Drives the Differences Across Estimands

The table above demonstrates a fundamental point: **the treatment effect has no single value without specifying the estimand**. The risk difference ranges from near zero to over 14 percentage points depending on how treatment switching is handled. This is not a statistical artifact---it reflects genuinely different causal questions about the same drug.

**Treatment-policy (RD > 0, moderate).** Under this estimand, the active drug increases risk because patients are followed regardless of switching. During the early period (before day 45 in the DGP), the active drug has HR = 1.25, causing excess harm. Some patients switch to the comparator, which reduces their subsequent hazard, but the early harm has already occurred. The treatment-policy RD captures the net real-world consequence of initiating the active drug: early harm partially offset by switching.

**Hypothetical no-switch (RD $\approx$ 0).** In a world where nobody switches, the early harm from the active drug (HR = 1.25 before day 45) is approximately offset by the late protective effect (HR = 0.70 after day 45). Over 180 days, these roughly cancel out. This estimand isolates the direct pharmacological effect of sustained treatment and is near null because the non-proportional hazard structure produces a time-averaged effect close to zero.

**While-on-treatment (RD $\approx$ 0).** The while-on-treatment truth equals the no-switch truth because, under counterfactual all-treated or all-control conditions, nobody switches (everyone is on the same drug). In observed data, however, the while-on-treatment estimand differs from the hypothetical because it conditions on not switching rather than intervening to prevent it---a distinction that matters when switching is informative.

**Composite (RD >> 0, largest).** The composite estimand absorbs switching into the outcome. Because the active drug causes more switching (driven by `gamma_A = 0.80`), the composite event rate is substantially higher in the active arm. This large risk difference is driven primarily by differential switching rates, not by differential renal failure rates. The composite estimand conflates drug tolerability with drug safety---a clinically important distinction that is lost when switching is treated as part of the outcome.

**Principal stratum (RD $\approx$ 0, smallest).** Among never-switchers---the approximately 19% of patients who would not switch under either treatment---the risk difference is near zero. Never-switchers tend to be lower-risk patients (without CKD) for whom the drug's effect is minimal. This estimand targets a highly selected, latent subpopulation that may not be representative of the patients clinicians are most concerned about.

::: {.callout-note}
## The same drug, five different answers
The active drug is genuinely riskier in the short term (HR = 1.25 early) and genuinely protective in the longer term (HR = 0.70 late). Whether the overall effect appears harmful, null, or protective depends entirely on which estimand is chosen---and each estimand answers a different clinical question. This is not a flaw of the framework; it is the framework working as intended. The flaw arises when analysts report a treatment effect without specifying which estimand they are targeting.
:::

## Data Support Diagnostics {#sec-support}

Identification of a causal estimand may be conceptually justified, but estimation can fail in practice if the data do not provide adequate empirical support for the target intervention. This section introduces diagnostics for assessing data support and demonstrates how estimation degrades as support weakens.

**Positivity** requires that, within every stratum defined by measured confounders, there is a nonzero probability of receiving each treatment. In longitudinal settings (as with LMTP), sequential positivity requires that the cumulative product of conditional treatment and censoring probabilities remains bounded away from zero at every time point. When this product approaches zero in some strata, the corresponding inverse-probability weights become extreme and estimators become unstable.

Rather than using fixed thresholds (e.g., flagging propensity scores below 0.05 or above 0.95), a data-adaptive approach sets the positivity bound as $5 / (\sqrt{n} \cdot \log n)$, where $n$ is the sample size (Gruber et al. 2022). For $n = 2000$, this gives bounds of approximately $(0.025, 0.975)$; for $n = 500$, approximately $(0.036, 0.964)$. The `compute_ps_diagnostics()` function in our support diagnostics reports both conventional and adaptive thresholds.

### Three Support Scenarios

We create three versions of the DGP by modifying parameters that control switching intensity and confounding.

```{r}
#| label: support-scenarios
#| cache: true

set.seed(101)
dat_good <- generate_hep_data(
  N = 5000, np_hazard = TRUE, dep_censor = TRUE,
  complexity = TRUE, policy = "treatment_policy", seed = 101
)

set.seed(102)
dat_strained <- generate_hep_data(
  N = 5000, np_hazard = TRUE, dep_censor = TRUE,
  complexity = TRUE, policy = "treatment_policy",
  gamma_A = 1.5, gamma_ckd = 1.2, lambda_sw0 = 5e-3,
  seed = 102
)

set.seed(103)
dat_poor <- generate_hep_data(
  N = 5000, np_hazard = TRUE, dep_censor = TRUE,
  complexity = TRUE, policy = "treatment_policy",
  gamma_A = 2.5, gamma_ckd = 2.0, lambda_sw0 = 1e-2,
  seed = 103
)
```

### Switching Probability Diagnostics

The three scenarios differ in switching parameters, not in treatment assignment. The relevant diagnostic is therefore the distribution of **predicted switching probabilities**, not treatment propensity scores (which are identical across scenarios because the treatment assignment model does not change).

```{r}
#| label: fig-support-switch
#| fig-cap: "Data support diagnostic: predicted switching probability by treatment arm across three support scenarios. As switching intensity increases, the active arm's switching probability distribution shifts rightward, creating informative censoring that biases conventional analyses."
#| fig-height: 9
#| fig-width: 7

scenario_list <- list(
  "1. Good support" = dat_good,
  "2. Strained support" = dat_strained,
  "3. Poor support" = dat_poor
)

support_plots <- plot_support_scenarios(scenario_list)
support_plots$by_treatment
```

```{r}
#| label: fig-support-ckd
#| fig-cap: "Switching probability by CKD status across scenarios. Under poor support, CKD patients have near-certain switching, violating the positivity assumption for estimands that require modelling the switching mechanism."
#| fig-height: 9
#| fig-width: 7

support_plots$by_ckd
```

### Support Summary Table

```{r}
#| label: tbl-support
#| tbl-cap: "Support diagnostic summary across scenarios"

support_tbl <- bind_rows(
  summarize_support(dat_good, "Good support"),
  summarize_support(dat_strained, "Strained support"),
  summarize_support(dat_poor, "Poor support")
)

kable(support_tbl, digits = 3, format = "pipe")
```

### How Estimation Degrades with Weak Support

```{r}
#| label: support-estimation
#| message: false
#| warning: false

# Load pre-computed support results (run R/run_cached_analyses.R to generate)
support_cache <- here("results", "support_estimation.rds")
stopifnot("Run R/run_cached_analyses.R first" = file.exists(support_cache))
support_results <- readRDS(support_cache)

kable(support_results, digits = 4, format = "pipe")
```

Under good support (12% switching), all methods produce estimates. As support degrades:

- The switching probability distribution shifts dramatically: under poor support, ~67% of patients switch and CKD patients switch near-deterministically
- Cox censor-at-switch HR diverges from naive (2.27 vs 1.39 under poor support) because censoring at switch removes a large, non-random fraction of the active arm
- LMTP estimates show increasing negative bias and wider confidence intervals as the censoring model must extrapolate further
- The practical lesson: identification may be conceptually valid while estimation fails due to lack of empirical support for the switching adjustment

::: {.callout-warning}
## Identification vs. estimation
A target estimand can be well-defined and conceptually identified under the assumed causal model, but estimation can still fail if the data do not contain enough subjects who follow (or could plausibly follow) the target intervention. Always check support diagnostics before interpreting estimates.
:::

## Repeated Simulation Study

We now assess estimator performance over repeated samples from the DGP. This reveals bias, variance, and coverage properties that a single dataset cannot show. For each estimand, we compare misaligned Cox estimators to the aligned LMTP estimator.

<!-- TODO(Joy): fill in exact runtime notes after benchmarking -->
<!-- TODO(Joy): benchmark whether 200 iterations is feasible or reduce for development -->

```{r}
#| label: run-sim
#| message: false
#| warning: false

source(here("R", "run_simulations.R"))

# Load pre-computed simulation results (run R/run_cached_analyses.R to generate)
sim_cache <- here("results", "sim_study_main.rds")
stopifnot("Run R/run_cached_analyses.R first" = file.exists(sim_cache))
sim_results <- readRDS(sim_cache)
```

### Results by Estimand

```{r}
#| label: tbl-sim-all
#| tbl-cap: "Simulation results by estimand (20 iterations, N = 10,000)"

estimand_labels <- c(
  treatment_policy   = "Treatment-policy",
  no_switch          = "Hypothetical no-switch",
  while_on_treatment = "While-on-treatment",
  composite          = "Composite",
  principal_stratum  = "Principal stratum"
)

sim_summaries <- lapply(names(truth_all), function(est) {
  if (!est %in% sim_results$estimand) return(NULL)
  sub <- sim_results %>% filter(estimand == est)
  s <- summarize_simulation(sub,
         truth_rd = truth_all[[est]]$true_rd,
         truth_rr = truth_all[[est]]$true_rr,
         truth_hr = truth_all[[est]]$true_hr)
  s$estimand_label <- estimand_labels[est]
  s
}) %>% bind_rows()

kable(sim_summaries %>%
        select(estimand_label, method, n_converged,
               mean_hr, bias_hr, mean_rd, bias_rd,
               mean_rr, bias_rr, coverage_rd) %>%
        rename(Estimand = estimand_label, Method = method),
      digits = 4, format = "pipe")
```

### Performance Plot

```{r}
#| label: fig-sim-performance
#| fig-cap: "Distribution of estimates across simulation iterations, by estimand and method. Red dashed line = true risk difference."
#| fig-height: 6
#| fig-width: 9

sim_plot_data <- sim_results %>%
  filter(converged) %>%
  mutate(
    estimate = case_when(
      method == "LMTP SDR" ~ risk_diff,
      TRUE ~ hr
    ),
    scale = case_when(
      method == "LMTP SDR" ~ "Risk Difference",
      TRUE ~ "Hazard Ratio"
    ),
    estimand_label = estimand_labels[estimand]
  )

truth_lines <- tibble(
  estimand_label = unname(estimand_labels[names(truth_all)]),
  scale = "Risk Difference",
  xint = sapply(truth_all, function(x) x$true_rd)
)

ggplot(sim_plot_data, aes(x = estimate, fill = method)) +
  geom_histogram(alpha = 0.6, bins = 25, position = "identity") +
  facet_grid(estimand_label ~ scale, scales = "free") +
  geom_vline(data = truth_lines,
             aes(xintercept = xint), linetype = "dashed", color = "red") +
  labs(x = "Estimate", y = "Count", fill = "Method",
       title = "Estimator Performance Within Each Estimand") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

```{r}
#| label: sim-convergence
conv_summary <- sim_results %>%
  group_by(estimand, method) %>%
  summarise(
    total = n(),
    converged = sum(converged),
    failed = sum(!converged),
    .groups = "drop"
  )
cat("Convergence summary:\n")
print(conv_summary)
```

## Interpretation

The simulation results reveal several important patterns about estimand-estimator alignment.

### Cox regression is dominated by confounding bias

Across all five estimands, the unadjusted Cox hazard ratios and Kaplan-Meier risk differences are severely biased. The KM risk difference averages approximately +0.08 when the true risk difference is near zero for most estimands, and Cox coverage is 0% in every case. This bias arises because CKD patients are more likely to receive the active drug *and* to experience renal failure. The confounding inflates the apparent treatment effect by a factor of 10--80 depending on the estimand.

The Cox naive and censor-at-switch estimators differ for the treatment-policy estimand (HR 1.21 vs 1.19; RD 0.086 vs 0.076), demonstrating that censoring at switch attenuates the estimate slightly by removing the highest-risk person-time. Under no-switch, the two are identical---as expected, since there is no switching to censor.

For the composite estimand, Cox performs relatively better (HR bias of +0.10 vs true HR of 1.70) because the large composite effect (driven by differential switching) dominates the confounding bias.

### LMTP reduces but does not eliminate bias

LMTP produces substantially less biased estimates than Cox for most estimands. For the no-switch and while-on-treatment estimands (where the true RD is near zero), LMTP achieves 75--85% coverage despite a residual negative bias of approximately -0.014. This residual bias likely reflects the limitations of the current specification: biweekly time bins (13 time points) provide limited temporal resolution for the non-proportional hazard structure, and the Super Learner library (`SL.mean`, `SL.glm`, `SL.bayesglm`) may not fully capture the confounding adjustment.

For the treatment-policy estimand (true RD = +0.007), LMTP coverage is 35%, reflecting a negative bias of -0.028. The estimator is correctly targeting the right parameter but the finite-sample performance is limited by the sparse event rate (~8%) and coarse time discretisation.

### The composite LMTP is misspecified

LMTP produces a catastrophically wrong estimate for the composite estimand (mean RD = -0.16 vs truth = +0.15). This occurs because the LMTP analysis targets static baseline treatment interventions, while the composite outcome includes switching as an event. The LMTP censoring model treats switching as a censoring event to be adjusted away, which is the *opposite* of what the composite estimand requires. This illustrates that even an aligned estimator framework can produce wrong answers when the intervention definition does not match the estimand.

### Principal stratum estimation is approximate

The LMTP analysis restricted to observed non-switchers (55% coverage) provides a rough approximation to the principal stratum estimand. The approximation is imperfect because observed non-switchers under one treatment arm are not the same as the latent "never-switcher" stratum (which requires not switching under *both* treatment assignments). More sophisticated principal stratum methods (e.g., sensitivity analysis under monotonicity) are needed for reliable estimation.

### The central lesson

The results confirm that the primary problem in safety analyses with intercurrent events is **estimand-estimator misalignment**. Cox regression, regardless of how switching is handled, produces severely biased estimates due to uncontrolled confounding and conditional (rather than marginal) targeting. LMTP offers better alignment with the marginal risk contrasts defined by most estimands, but its performance depends on adequate temporal resolution, a sufficiently rich Super Learner library, and correct specification of the intervention relative to the estimand. Neither method is universally "better"---the question is whether the estimator targets the same quantity as the estimand.

## Sensitivity to Unmeasured Confounding

Even with a well-aligned estimator and adequate data support, the causal interpretation of any observational estimate rests on the assumption of no unmeasured confounding (exchangeability). The **causal gap** $\delta = \psi_{\text{stat}} - \psi_{\text{causal}}$ quantifies the difference between what the statistical estimand identifies and the true causal quantity. In a randomised trial, $\delta = 0$ by design; in an observational study, $\delta$ reflects residual confounding from unmeasured common causes of treatment and outcome.

A useful diagnostic is the **G-value** (Gruber et al. 2024): the minimum causal gap needed to shift the confidence interval to include the null. If the G-value is large relative to the magnitude of the measured confounding adjustment (the difference between unadjusted and adjusted estimates), the finding is robust. If it is small, even modest unmeasured confounding could nullify the result. Sensitivity analyses should examine how the point estimate and confidence interval change across a range of hypothesised $\delta$ values, typically from 0 to 5 times the measured confounding adjustment (Dang et al. 2023).

This tutorial does not implement a full sensitivity analysis, but the causal roadmap framework requires it as a distinct step (Step 6). In a complete analysis, a sensitivity plot showing shifted estimates and confidence bounds across a range of causal gap values should accompany every primary result.

<!-- TODO(Joy): implement a nonparametric sensitivity analysis function and
     add a worked example to the simulation study. -->

## Key Messages

1. **The same drug, five different answers.** The true risk difference ranges from near zero (hypothetical no-switch, principal stratum) to +0.15 (composite), depending entirely on how treatment switching is handled. The treatment effect has no single value without specifying the estimand.

2. **Cox regression is dominated by confounding bias.** Across all five estimands, unadjusted Cox models produce 0% coverage. The confounding from CKD (which drives both treatment assignment and the outcome) inflates the apparent treatment effect by an order of magnitude. This bias is not a property of the Cox model per se, but of using unadjusted or insufficiently adjusted regression in the presence of strong confounding.

3. **Hazard ratios are not risk contrasts.** Even when the Cox HR is close to the true HR (as in the composite estimand), the KM risk difference remains biased because unadjusted survival curves do not target the marginal risk under intervention. Estimands defined as risk contrasts require estimators that target risk contrasts.

4. **LMTP reduces bias but does not eliminate it.** LMTP achieves 75--85% coverage for the no-switch and while-on-treatment estimands (near-null true effects) but only 35% for the treatment-policy estimand. Performance depends on temporal resolution, Super Learner specification, and event rate. Production analyses should use finer time bins and richer learner libraries.

5. **Misspecified interventions produce wrong answers.** The composite LMTP fails catastrophically (bias = -0.31) because the LMTP intervention (static treatment assignment) does not match the composite estimand (which absorbs switching into the outcome). Even a well-designed estimator framework produces wrong answers when the intervention definition is misaligned with the estimand.

6. **Support diagnostics are mandatory.** Estimation degrades as positivity weakens. Under poor support, Cox HRs become unstable (HR = 2.27 for censor-at-switch) and LMTP estimates show increased bias and wider confidence intervals.

7. **Report transparently.** State the estimand, the assumptions required for identification, the diagnostics that support (or fail to support) those assumptions, the alignment between the estimator and the estimand, and the limitations of the specification (time binning, learner library, finite-sample bias).

## Sources

- ICH E9(R1) Addendum (2019). Addendum on estimands and sensitivity analyses in clinical trials.
- Williams N, Diaz I (2023). lmtp: Non-parametric causal effects of feasible interventions based on modified treatment policies. R package.
- van der Laan MJ, Rose S (2011). Targeted Learning: Causal Inference for Observational and Experimental Data.
- Hernan MA, Robins JM (2020). Causal Inference: What If.
- Dang LE, Tarp JM, Abrahamsen B, et al. (2023). A causal roadmap for generating high-quality real-world evidence. Journal of Clinical and Translational Science.
- Gruber S, Lee H, Phillips R, Ho M, van der Laan MJ (2022). Developing a targeted learning-based statistical analysis plan. Statistics in Biopharmaceutical Research.
- Gruber S, Phillips RV, Lee H, van der Laan MJ (2024). Targeted learning: toward a future informed by real-world evidence. Statistics in Biopharmaceutical Research.
- Chen RC, Rosenblum M, van der Laan MJ, Gruber S (2023). Beyond the Cox hazard ratio: a targeted learning approach to survival analysis in a cardiovascular outcome trial application. Clinical Trials.
- Martinussen T (2022). Causality and the Cox regression model. Annual Review of Statistics and Its Application.
- Stensrud MJ, Aalen JM, Aalen OO, Valberg M (2019). Limitations of hazard ratios in clinical trials. European Heart Journal.
- Rufibach K (2019). Treatment effect quantification for time-to-event endpoints: estimands, analysis strategies, and beyond. Pharmaceutical Statistics.
- Petersen ML (2014). Applying a causal road map in settings with time-dependent confounding. Epidemiology.
- Rytgaard HC, Eriksson F, van der Laan MJ (2023). Estimation of time-specific intervention effects on continuously distributed time-to-event outcomes by targeted maximum likelihood estimation. Biometrics, 79(4):3038--3049.
- Chen D, Rytgaard HC, Fong E, Tarp JM, Petersen ML, van der Laan MJ, Gerds TA (2023). concrete: Targeted estimation of survival and competing risks in continuous time. arXiv:2310.19197.
