---
title: "Estimand–Estimator Alignment in Post-Authorization Safety Analyses"
subtitle: "A causal roadmap approach to intercurrent events in time-to-event studies"
author:
  - name: "Joy Nakato"
    affiliation: "University of California, Berkeley"
date: today
format:
  revealjs:
    theme: [default, custom.scss]
    slide-number: true
    width: 1280
    height: 720
    transition: fade
    fig-align: center
    code-fold: true
    echo: false
    message: false
    warning: false
    fig-dpi: 200
    fig-format: png
execute:
  cache: true
bibliography: []
---

```{r}
#| label: setup
#| include: false
#| cache: false

# ── Source files from the estimand_tutorial project ──
# ground_truth.rds, lmtp_main.rds, support_estimation.rds
# DGP.R, R/helpers.R, R/support_diagnostics.R

library(tidyverse)
library(survival)
library(here)
library(knitr)

# Presentation theme
theme_pres <- theme_minimal(base_size = 16) +
  theme(
    plot.title = element_text(size = 18, face = "bold"),
    strip.text = element_text(size = 14),
    legend.position = "bottom",
    legend.text = element_text(size = 12),
    panel.grid.minor = element_blank()
  )
theme_set(theme_pres)

pal <- c("steelblue", "tomato", "seagreen", "darkorange", "mediumpurple")

# Source project code
source(here("DGP.R"))
source(here("R", "helpers.R"))

# Load cached results
truth_cache <- here("results", "sim_results", "ground_truth.rds")
lmtp_cache  <- here("results", "lmtp_main.rds")
sup_cache   <- here("results", "support_estimation.rds")

has_truth <- file.exists(truth_cache)
has_lmtp  <- file.exists(lmtp_cache)
has_sup   <- file.exists(sup_cache)

if (has_truth) truth_all <- readRDS(truth_cache)
if (has_lmtp) lmtp_res  <- readRDS(lmtp_cache)
if (has_sup)  sup_res    <- readRDS(sup_cache)
```

## The problem {.smaller}

In post-authorization safety studies, the question *"does this drug increase the risk of [adverse event]?"* has no single answer until several decisions are made:

. . .

:::: {.columns}
::: {.column width="55%"}
**Intercurrent events complicate the question:**

- Treatment switching or discontinuation
- Informative monitoring and censoring
- Competing events (e.g., death)
- Loss to follow-up dependent on health status

Each changes what the analysis is actually estimating.
:::

::: {.column width="45%"}
**Three distinct failure modes:**

1. **Wrong estimand** — the analysis answers a different question than intended
2. **Right estimand, wrong estimator** — the method does not target the stated quantity
3. **Right estimand and estimator, poor support** — data lack positivity for the target intervention
:::
::::

::: {.notes}
Frame the talk: this is not about a specific drug. It is about the analytical choices that determine what any post-marketing safety analysis is actually estimating.
:::

## Running example: antiviral nephrotoxicity {.smaller}

:::: {.columns}
::: {.column width="60%"}

**Setting:** Post-marketing comparison of two hepatitis B antivirals for renal safety

- **Active:** tenofovir-containing regimen (TDF)
- **Comparator:** entecavir-based regimen (ETV)
- **Outcome:** renal failure within 180 days
- **Intercurrent event:** treatment switching driven by early renal signals

**Key complication:** CKD drives both switching and the outcome — switching is informative.
:::

::: {.column width="40%"}

**Published evidence:**

- TDF associated with modestly elevated early renal risk vs ETV (Lee et al. 2019)
- Risk higher among patients with baseline CKD (Jung et al. 2022; Mak et al. 2022)
- Effect attenuates over longer follow-up
:::
::::

::: {.notes}
This is a simulated example calibrated to the published nephrotoxicity literature. The DGP uses HR=2.0 in the first 90 days, HR=0.90 after, with informative switching (~11% by 180 days).
:::

## The causal roadmap {.smaller}

The analysis should proceed through a structured sequence, not jump directly to a regression model.

```{r}
#| label: fig-roadmap
#| fig-width: 11
#| fig-height: 3.5

library(ggplot2)

steps <- tibble(
  step = factor(1:6),
  label = c("1. Causal\nquestion", "2. Causal\nmodel (DAG)",
            "3. Identifiability\nassessment",
            "4. Statistical\nestimand",
            "5. Estimator\nchoice",
            "6. Sensitivity\nanalysis"),
  x = seq(1, 6),
  y = 0
)

ggplot(steps, aes(x = x, y = y)) +
  geom_segment(aes(x = x + 0.35, xend = x + 0.65, y = 0, yend = 0),
               arrow = arrow(length = unit(0.15, "cm"), type = "closed"),
               color = "grey40", linewidth = 0.8,
               data = steps %>% filter(step != 6)) +
  geom_label(aes(label = label), size = 4.5, fill = "white",
             label.size = 0.5, label.padding = unit(0.4, "lines"),
             fontface = "bold") +
  annotate("text", x = 3.5, y = -0.25,
           label = "Dang et al. 2023; Petersen 2014",
           size = 3.5, color = "grey50", fontface = "italic") +
  coord_cartesian(xlim = c(0.3, 6.7), ylim = c(-0.4, 0.4)) +
  theme_void(base_size = 14)
```

. . .

> Most analyses jump from step 1 to step 5. Steps 2–4 are where estimand–estimator mismatches originate.

::: {.notes}
The causal roadmap (Dang et al. 2023, Petersen 2014) organizes the analysis into a principled sequence. Most conventional analyses skip the middle steps.
:::

## Causal structure {.smaller}

```{r}
#| label: fig-dag
#| fig-width: 10
#| fig-height: 5.5

library(dagitty)
library(ggdag)

dag <- dagify(
  Treatment ~ W,
  Switch ~ Treatment + W + Kidney,
  Kidney ~ Treatment + W,
  Censoring ~ Switch + W + Kidney,
  AKI ~ Treatment + Switch + W + Kidney,
  exposure = "Treatment",
  outcome = "AKI",
  labels = c(
    W = "Baseline\nconfounders\n(CKD, age, ...)",
    Treatment = "Treatment\nassignment\n(TDF vs ETV)",
    Kidney = "Evolving\nkidney\nfunction",
    Switch = "Treatment\nswitch",
    Censoring = "Censoring /\nmonitoring",
    AKI = "Renal\nfailure"
  ),
  coords = list(
    x = c(W = 0, Treatment = -1, Kidney = 1, Switch = 1, Censoring = 2.5, AKI = 2),
    y = c(W = 1, Treatment = 0, Kidney = 0, Switch = -0.8, Censoring = -0.8, AKI = 0.5)
  )
)

ggdag(dag, text = FALSE, use_labels = "label", text_size = 4,
      node_size = 20, stylized = FALSE) +
  theme_dag(base_size = 14) +
  theme(plot.margin = margin(5, 5, 5, 5))
```

::: {.notes}
Key: W (especially CKD) is a common cause of treatment assignment, switching, kidney function, and AKI. Switching is driven by evolving kidney function, making it informative. Conditioning on post-treatment variables (switching, censoring) without adjustment can open collider paths.
:::

## Five estimands, one dataset {.smaller}

The ICH E9(R1) addendum defines five strategies for handling the same intercurrent event. Each defines a different causal quantity — even from the same data.

::: {.incremental}

**1. Treatment-policy** | Follow all patients regardless of switching | *Real-world effect of initiating treatment, including downstream switching*

**2. Hypothetical no-switch** | Model / intervene to prevent switching | *Effect if no patient could switch — isolates the direct pharmacological effect*

**3. While-on-treatment** | Switching treated as a censoring event | *Effect of the drug before switching — 180-day risk with follow-up truncated at switch; L-TMLE censoring model adjusts for informative switching*

**4. Composite** | Absorb switching into the outcome | *Risk of renal failure OR switching — conflates tolerability with safety*

**5. Principal stratum** | Restrict to latent never-switchers | *Effect among patients who would not switch under **either** drug — a strong cross-world assumption; only observable approximation is non-switchers under their observed treatment*

:::

. . .

Each answers a **different clinical question**. Choosing among them is a scientific decision, not a statistical one.

::: {.notes}
All five are implemented in the simulation with distinct L-TMLE specifications: baseline-only treatment (treatment-policy, composite, principal stratum), time-varying hold-constant (no-switch), and competing-risks (while-on-treatment). Principal stratum uses an oracle never-switcher indicator in simulation.
:::

## True effects differ across estimands {.smaller}

Under the simulation DGP, the same drug produces different relative risks depending on the estimand strategy.

```{r}
#| label: fig-truth
#| fig-width: 9
#| fig-height: 7.5

if (has_truth) {
  truth_df <- tibble(
    Strategy = c("Treatment-\npolicy", "No-switch\n(hypothetical)",
                 "While-on-\ntreatment", "Composite", "Principal\nstratum"),
    RR = c(truth_all$treatment_policy$true_rr,
           truth_all$no_switch$true_rr,
           truth_all$while_on_treatment$true_rr,
           truth_all$composite$true_rr,
           truth_all$principal_stratum$true_rr),
    Strategy_order = 1:5
  ) %>%
    mutate(Strategy = fct_reorder(Strategy, Strategy_order))

  estimand_pal <- c(
    "Treatment-\npolicy"        = "#1E2761",
    "No-switch\n(hypothetical)" = "#028090",
    "While-on-\ntreatment"      = "#2C5F2D",
    "Composite"                 = "#B85042",
    "Principal\nstratum"        = "#6D2E46"
  )

  ggplot(truth_df, aes(x = Strategy, y = RR, color = Strategy)) +
    geom_hline(yintercept = 1, linetype = "dashed", color = "grey40") +
    geom_point(size = 10) +
    geom_text(aes(label = sprintf("%.2f", RR)),
              vjust = -1.5, size = 5, fontface = "bold", show.legend = FALSE) +
    scale_color_manual(values = estimand_pal, guide = "none") +
    labs(x = NULL, y = "True 180-day relative risk",
         title = "True RR by ICE strategy") +
    coord_cartesian(ylim = c(0.95, max(truth_df$RR) * 1.12)) +
    theme(axis.text.x = element_text(size = 18, face = "bold"),
          axis.text.y = element_text(size = 14),
          axis.title.y = element_text(size = 16))
} else {
  plot.new()
  text(0.5, 0.5, "Run R/run_cached_analyses.R to generate truth", cex = 1.5)
}
```

::: {.notes}
Strategies are ordered to match the headline coverage table. The composite RR (1.46) is the largest because differential switching rates inflate the composite endpoint. Treatment-policy (1.25), no-switch (1.34), while-on-treatment (1.26), and principal stratum (1.35) differ because each defines a different counterfactual contrast. This relative ordering is DGP-specific — under different switching dynamics the WOT effect can be the largest.
:::


## Why Cox regression may not target your estimand {.smaller}

:::: {.columns}
::: {.column width="55%"}

**The Cox hazard ratio is a conditional parameter.** It does not generally equal the marginal risk contrast defined by any of the estimands above.

Three issues compound:

1. **Noncollapsibility:** The marginal HR differs from the conditional HR even without confounding (Martinussen 2022)
2. **Built-in selection:** At later time points, high-risk subjects are differentially depleted (Stensrud et al. 2019)
3. **Informative censoring:** Censoring at switch removes the highest-risk person-time, changing the risk set composition
:::

::: {.column width="45%"}
```{r}
#| label: fig-cox-riskset
#| fig-width: 3.5
#| fig-height: 3
#| out-width: "75%"

# Illustrate how censoring at switch distorts the risk set
set.seed(42)
n <- 200
risk <- rbeta(n, 2, 5)
trt <- rbinom(n, 1, 0.5)
switch_prob <- plogis(-2 + 3 * risk + 0.5 * trt)
switched <- rbinom(n, 1, switch_prob)

df_riskset <- tibble(
  `Baseline risk` = risk,
  Treatment = factor(trt, labels = c("Comparator", "Active")),
  Status = ifelse(switched == 1, "Switched (censored)", "Remained")
)

ggplot(df_riskset %>% filter(Treatment == "Active"),
       aes(x = `Baseline risk`, fill = Status)) +
  geom_histogram(bins = 25, alpha = 0.7, position = "identity") +
  scale_fill_manual(values = c("Remained" = "steelblue",
                                "Switched (censored)" = "tomato")) +
  labs(title = "Active arm risk profile",
       subtitle = "Censoring at switch removes highest-risk patients",
       y = "Count") +
  theme(legend.position = c(0.7, 0.8))
```
:::
::::

::: {.notes}
This slide makes the point visually: the patients who switch away are the sickest. Censoring them distorts the remaining risk set, making the drug appear safer than it is. This is not a general critique of Cox regression — it is a specific concern when censoring is informative.
:::


## Estimand-aligned estimation {.smaller}

The estimator should follow from the estimand, not the reverse.

:::: {.columns}
::: {.column width="50%"}

**Conventional approaches:**

| Analysis | What it estimates |
|---|---|
| Cox, ignore switch | Conditional HR; risk set altered by switching |
| Cox, censor at switch | Conditional HR among non-switchers; biased if switching informative |
| Cox, time-dependent trt | Conditional HR for current exposure; not a marginal target |
:::

::: {.column width="50%"}

**Estimand-aligned approaches:**

| Method | Target |
|---|---|
| G-computation | Marginal risk (requires correct outcome model) |
| IPTW | Marginal risk (requires correct weight model) |
| L-TMLE | Marginal risk under specified intervention |

All require specifying the intervention that defines the estimand.
:::
::::

. . .

> The SDR estimator is **doubly robust**: consistent if *either* the outcome model *or* the treatment/censoring model is correctly specified.

::: {.notes}
Key principle: the estimator must be chosen to target the parameter defined by the estimand. A Cox HR is always interpretable as a Cox HR — the question is whether that is the quantity the scientific question requires.
:::


## Simulation design {.smaller}

```{r}
#| label: fig-simdesign
#| fig-width: 10
#| fig-height: 4

# Timeline schematic
timeline <- tibble(
  phase = c("Early period\n(0–90 days)", "Late period\n(90–180 days)"),
  x = c(45, 135),
  hr = c("HR = 2.0", "HR = 0.90"),
  note = c("TDF increases\nrenal risk", "Effect attenuates,\nmodest reversal")
)

ggplot() +
  # Timeline bar
  annotate("rect", xmin = 0, xmax = 90, ymin = -0.15, ymax = 0.15,
           fill = "tomato", alpha = 0.3) +
  annotate("rect", xmin = 90, xmax = 180, ymin = -0.15, ymax = 0.15,
           fill = "steelblue", alpha = 0.3) +
  # Labels
  annotate("text", x = 45, y = 0.35, label = "HR = 2.0\n(early harm)",
           size = 5, fontface = "bold", color = "tomato") +
  annotate("text", x = 135, y = 0.35, label = "HR = 0.90\n(attenuated)",
           size = 5, fontface = "bold", color = "steelblue") +
  # Switching annotation
  annotate("segment", x = 60, xend = 60, y = -0.15, yend = -0.4,
           arrow = arrow(length = unit(0.2, "cm")), color = "darkorange") +
  annotate("text", x = 60, y = -0.55, label = "~11% switch\n(informative)",
           size = 4, color = "darkorange") +
  # Axis
  annotate("segment", x = 0, xend = 180, y = 0, yend = 0,
           linewidth = 1.2) +
  annotate("text", x = c(0, 90, 180), y = -0.08,
           label = c("Day 0", "Day 90", "Day 180"), size = 4) +
  # DGP notes
  annotate("text", x = 90, y = -0.85,
           label = "N = 10,000 | Confounding by CKD, cirrhosis, age | Non-proportional hazards",
           size = 3.5, color = "grey40") +
  coord_cartesian(xlim = c(-10, 190), ylim = c(-1, 0.6)) +
  theme_void(base_size = 14) +
  labs(title = "Data-generating process: hepatitis B renal safety")
```

::: {.notes}
The DGP uses HR=2.0 (early) and HR=0.90 (late) with ~16% event rate and ~11% switching. Treatment assignment is confounded by CKD, cirrhosis, and age. Switching is driven by treatment and CKD, making it informative. The non-proportional hazard structure means any single HR is a time-averaged summary.
:::


## Performance metrics {.smaller}

Each simulated dataset is analysed with four estimators. We summarise performance on two scales:

:::: {.columns}
::: {.column width="50%"}
**Estimand scale (risk difference)**

- True RD is known by Monte Carlo (N = 500,000 per arm)
- **Bias** = mean estimate − truth
- **RD coverage** = proportion of 95% CIs containing truth
  - L-TMLE: analytic via efficient influence function
  - Cox: bootstrap (200 replicates) on KM risk difference
:::

::: {.column width="50%"}
**Estimator scale (hazard ratio)**

- True marginal HR from a Cox regression on counterfactual potential outcomes
- **HR coverage** = proportion of Cox HR CIs containing the true marginal HR
- L-TMLE does not target an HR; not evaluated on this scale
:::
::::

. . .

Reporting both exposes distinct failure modes:

- **Wrong scale:** the Cox HR reports on a scale that does not correspond to the estimand
- **Wrong adjustment:** even on its own HR scale, Cox fails to recover the correct marginal HR when confounding or informative switching is present

::: {.notes}
Bias on the HR scale is sensitive to the non-collapsibility of the HR; we focus on coverage. The true marginal HR is computed via compute_true_hr in calc_truth.R, which fits coxph(Surv ~ trt) on the combined counterfactual all-treated and all-control datasets.
:::


## Single-dataset results {.smaller}

```{r}
#| label: tbl-single-dataset

# Load cached L-TMLE results (no live simulation in the presentation)
# Cox fits are fast enough to run inline
set.seed(2026)
dat <- generate_hep_data(N = 10000, h0 = 5e-3, HR_early = 2.0, HR_late = 0.90,
                         np_hazard = TRUE, dep_censor = TRUE, complexity = FALSE,
                         lambda_sw0 = 1e-3, gamma_A = 0.80, gamma_ckd = 0.60,
                         seed = 2026)
tau <- 180
covars <- c("age", "ckd", "cirrhosis", "diabetes", "heart_failure")

cox_n  <- fit_cox_naive(dat, tau = tau, covars = covars)
cox_cs <- fit_cox_censor_switch(dat, tau = tau, covars = covars)
km_n   <- km_risk_at(cox_n$survfit, t = tau)
km_cs  <- km_risk_at(cox_cs$survfit, t = tau)

results_single <- tibble(
  Analysis = c("Cox (baseline trt)", "Cox (censor at switch)", "L-TMLE"),
  Scale = c("HR", "HR", "RD"),
  Estimate = c(round(cox_n$hr, 3), round(cox_cs$hr, 3),
               if (has_lmtp) round(-lmtp_res$contrast_rd$estimates$estimate, 4) else NA),
  `KM RD` = c(round(km_n$risk_diff, 4), round(km_cs$risk_diff, 4),
              if (has_lmtp) round(-lmtp_res$contrast_rd$estimates$estimate, 4) else NA),
  `True TP RD` = rep(if (has_truth) round(truth_all$treatment_policy$true_rd, 4) else NA, 3)
)

kable(results_single, format = "pipe",
      caption = "Estimates from a single simulated dataset (N = 10,000)")
```

. . .

The Cox HR and unadjusted KM risk difference reflect **confounding** — CKD patients are more likely to receive TDF *and* to develop renal failure. The L-TMLE risk difference is closer to the marginal truth but subject to finite-sample variability.

::: {.notes}
Do not compare the Cox HR and L-TMLE RD as though they estimate the same quantity. The point is that they target different parameters, and the Cox analyses are further biased by confounding.
:::


## Repeated simulation results {.smaller}

```{r}
#| label: fig-sim-results
#| fig-width: 10
#| fig-height: 5
#| cache.extra: !expr file.info(here::here("results", "sim_study_main.rds"))$mtime

# Prefer production cache; fall back to test cache if production not yet complete
sim_cache <- if (file.exists(here("results", "sim_study_main.rds"))) {
  here("results", "sim_study_main.rds")
} else {
  here("results", "sim_study_test.rds")
}
has_sim <- file.exists(sim_cache)

if (has_sim && has_truth) {
  source(here("R", "run_simulations.R"))
  sim_results <- readRDS(sim_cache)

  estimand_labels <- c(
    treatment_policy = "Treatment-policy",
    no_switch = "No-switch"
  )

  est_keep <- c("treatment_policy", "no_switch")

  lmtp_rr <- sim_results %>%
    filter(converged, !is.na(risk_ratio),
           estimand %in% est_keep,
           method %in% c("LMTP SDR", "LMTP SDR (no-switch)")) %>%
    transmute(estimand_label = estimand_labels[estimand],
              Estimator = "L-TMLE",
              metric = "Relative risk",
              value  = risk_ratio)

  cox_hr <- sim_results %>%
    filter(converged, !is.na(hr),
           estimand %in% est_keep,
           method == "Cox naive") %>%
    transmute(estimand_label = estimand_labels[estimand],
              Estimator = "Cox (naive)",
              metric = "Hazard ratio",
              value  = hr)

  sim_plot <- bind_rows(lmtp_rr, cox_hr)

  truth_lines <- tibble(
    estimand_label = rep(c("Treatment-policy", "No-switch"), each = 2),
    Estimator = rep(c("L-TMLE", "Cox (naive)"), 2),
    xint = c(truth_all$treatment_policy$true_rr, truth_all$treatment_policy$true_hr,
             truth_all$no_switch$true_rr,        truth_all$no_switch$true_hr))

  n_iter_used <- length(unique(sim_results$iter[sim_results$method == "LMTP SDR"]))

  ggplot(sim_plot, aes(x = value, fill = Estimator, color = Estimator)) +
    geom_histogram(alpha = 0.55, bins = 25, position = "identity") +
    facet_wrap(~ estimand_label, scales = "free_x") +
    geom_vline(data = truth_lines,
               aes(xintercept = xint, color = Estimator),
               linetype = "dashed", linewidth = 1, show.legend = FALSE) +
    scale_fill_manual(values = c("L-TMLE" = "steelblue",
                                  "Cox (naive)" = "tomato")) +
    scale_color_manual(values = c("L-TMLE" = "steelblue",
                                   "Cox (naive)" = "tomato")) +
    labs(x = "Estimate (L-TMLE: marginal RR; Cox: marginal HR)",
         y = "Count",
         title = sprintf("L-TMLE RR vs. Cox HR estimates across %d iterations (N = 10,000 each)",
                         n_iter_used),
         subtitle = "Dashed lines: true marginal RR (blue) and true marginal HR (red)") +
    theme(legend.position = "bottom")
} else {
  # Placeholder if simulation hasn't been run yet
  ggplot() +
    annotate("text", x = 0.5, y = 0.5,
             label = "Simulation results pending\n\nRun source('R/run_cached_analyses.R')\nto generate sim_study_main.rds",
             size = 6, color = "grey50") +
    theme_void() +
    labs(title = "Repeated simulation results")
}
```

::: {.notes}
This slide loads from cached simulation results (sim_study_main.rds). Production target is 500 iterations with an expanded Super Learner library (mean / glm / bayesglm / glmnet / ranger); the rds will refresh when the rerun completes. The simulation compares Cox naive, Cox censor-at-switch, Cox IPCW, and estimand-specific L-TMLE (SDR) fits. DGP: HR_early=2.0, HR_late=0.90, ~11% switching, ~16% event rate, ~19% never-switchers.
:::


## Headline findings: coverage by estimator and estimand {.smaller}

```{r}
#| label: tbl-headline-coverage
#| cache.extra: !expr file.info(here::here("results", "sim_study_main.rds"))$mtime

sim_cache <- here("results", "sim_study_main.rds")
truth_cache_full <- here("results", "sim_results", "ground_truth.rds")

if (file.exists(sim_cache) && file.exists(truth_cache_full)) {
  sim <- readRDS(sim_cache)
  truth <- readRDS(truth_cache_full)
  trd <- sapply(truth, function(x) x$true_rd)
  thr <- sapply(truth, function(x) x$true_hr)
  trr <- sapply(truth, function(x) x$true_rr)

  est_order <- c("treatment_policy", "no_switch", "while_on_treatment",
                 "composite", "principal_stratum")
  est_label <- c(treatment_policy = "Treatment-policy",
                 no_switch = "No-switch",
                 while_on_treatment = "While-on-treatment",
                 composite = "Composite",
                 principal_stratum = "Principal stratum")

  lmtp_match <- c(treatment_policy = "LMTP SDR",
                  no_switch = "LMTP SDR (no-switch)",
                  while_on_treatment = "LMTP SDR (WOT)",
                  composite = "LMTP SDR (composite)",
                  principal_stratum = "LMTP SDR (oracle PS)")

  cox_match <- c(treatment_policy   = "Cox naive",
                 no_switch          = "Cox IPCW",
                 while_on_treatment = "Cox censor-at-switch",
                 composite          = "Cox naive",
                 principal_stratum  = "Cox naive")

  cox_short <- c("Cox naive" = "naive on full follow-up",
                 "Cox IPCW"  = "IPCW for switching",
                 "Cox censor-at-switch" = "censor at switch",
                 "Cox naive (composite)" = "naive on composite endpoint")

  fmtp <- function(x) ifelse(is.na(x), "—", sprintf("%.0f%%", 100 * x))

  rows <- lapply(est_order, function(e) {
    lmtp_e <- sim %>% filter(method == lmtp_match[e], estimand == e, converged)
    cox_e  <- sim %>% filter(method == cox_match[e], estimand == e, converged)

    rd_cov  <- mean(lmtp_e$ci_low <= trd[e] & lmtp_e$ci_high >= trd[e],
                    na.rm = TRUE)
    rr_cov  <- mean(lmtp_e$rr_ci_low <= trr[e] & lmtp_e$rr_ci_high >= trr[e],
                    na.rm = TRUE)
    hr_cov  <- mean(cox_e$hr_ci_low <= thr[e] & cox_e$hr_ci_high >= thr[e],
                    na.rm = TRUE)

    tibble(
      Estimand   = est_label[e],
      `True RR`  = sprintf("%.2f",  trr[e]),
      `True HR`  = sprintf("%.2f",  thr[e]),
      `L-TMLE RD coverage` = fmtp(rd_cov),
      `L-TMLE RR coverage` = fmtp(rr_cov),
      `Aligned Cox HR coverage` =
        sprintf("%s (%s)", fmtp(hr_cov), cox_short[cox_match[e]])
    )
  })

  knitr::kable(bind_rows(rows), format = "pipe",
               caption = sprintf("Coverage of 95%% CIs across %d simulated datasets (N = 10,000 each)",
                                 length(unique(sim$iter))))
} else {
  cat("*Simulation results not yet cached.*")
}
```

. . .

Two findings deserve emphasis:

- **L-TMLE achieves near-nominal coverage on both the RD and RR scales** for every estimand it directly targets — including while-on-treatment once treatment is included as a (discretized) time-varying node in the L-TMLE fit.
- **Cox HR confidence intervals confidently exclude the true marginal HR.** For the treatment-policy estimand HR coverage is effectively 0%; the only setting where Cox HR coverage approaches nominal is the principal-stratum oracle, where the never-switcher subgroup happens to resemble the overall cohort under this DGP.

::: {.notes}
The HR coverage finding is the most striking quantitative result. "0% coverage" is not a wide-CI-miss but a confident wrong answer. The PS-via-Cox finding is a teaching point — under stronger informative switching the Cox HR would also miss on the principal stratum. The relative ordering of true RDs across estimands is DGP-specific: in other scenarios the while-on-treatment effect can be the largest. WOT performance improved markedly after reformulating L-TMLE to absorb switching as a competing event in the outcome (`competing_risk_at_switch = TRUE`).
:::


## Extension to other intercurrent events {.smaller}

::: {style="font-size: 0.75em;"}

The estimand framework applies to any post-baseline event, not just treatment switching.

:::: {.columns}
::: {.column width="50%"}
**Informative loss to follow-up**

- Treatment-policy: follow regardless of disenrollment
- Hypothetical: model a world without dropout
- WOT: censor at disenrollment (needs censoring model)

*L-TMLE C columns handle this structurally identical to switch-censoring.*
:::

::: {.column width="50%"}
**Informative monitoring**

- More-monitored patients have more chances for events to be detected
- L-TMLE models monitoring as a time-varying covariate jointly with outcome

**Competing risks**

- Death from unrelated causes precludes the safety outcome
- Specify strategy: censor, composite, or cause-specific (e.g., `concrete` R package for continuous-time TMLE)
:::
::::

. . .

**The same three lessons apply:**

1. The Cox HR does not generally correspond to the marginal risk contrast defined by the estimand.
2. Conventional Cox-based adjustments (censor, reweight) can amplify bias when the adjustment model is misspecified.
3. Estimand-aligned approaches (L-TMLE, TMLE) target the marginal quantity directly but depend on positivity and model specification.

:::

::: {.notes}
The goal of this slide is to generalize beyond the hepatitis B switching scenario. The pattern of findings — wrong scale, wrong adjustment, L-TMLE as a structural fix — applies to other intercurrent events with the same logic.
:::


## Positivity and support {.smaller}

Even an aligned estimator fails when the data do not support the target intervention.

```{r}
#| label: fig-support
#| fig-width: 10
#| fig-height: 5

# Small datasets for the switching probability illustration (fast, N=2000)
set.seed(101)
dat_good_sm <- generate_hep_data(N = 2000, h0 = 5e-3, HR_early = 2.0, HR_late = 0.90,
                                  np_hazard = TRUE, dep_censor = TRUE, complexity = FALSE,
                                  lambda_sw0 = 1e-3, gamma_A = 0.80, gamma_ckd = 0.60,
                                  seed = 101)
set.seed(103)
dat_poor_sm <- generate_hep_data(N = 2000, h0 = 5e-3, HR_early = 2.0, HR_late = 0.90,
                                  np_hazard = TRUE, dep_censor = TRUE, complexity = FALSE,
                                  gamma_A = 2.0, gamma_ckd = 1.5,
                                  lambda_sw0 = 8e-3, seed = 103)

sw_good <- glm(switched ~ age + ckd + treatment + cirrhosis,
               data = dat_good_sm, family = binomial)
sw_poor <- glm(switched ~ age + ckd + treatment + cirrhosis,
               data = dat_poor_sm, family = binomial)

combined <- bind_rows(
  dat_good_sm %>% mutate(p_switch = predict(sw_good, type = "response"),
                          Scenario = "Good support (~11% switching)"),
  dat_poor_sm %>% mutate(p_switch = predict(sw_poor, type = "response"),
                          Scenario = "Poor support (~50% switching)")
) %>%
  mutate(trt_label = ifelse(treatment == 1, "Active (TDF)", "Comparator (ETV)"))

ggplot(combined, aes(x = p_switch, fill = trt_label)) +
  geom_histogram(alpha = 0.6, bins = 40, position = "identity") +
  facet_wrap(~ Scenario, scales = "free_y", ncol = 2) +
  scale_fill_manual(values = c("tomato", "steelblue")) +
  labs(x = "Predicted P(switch | baseline covariates)",
       y = "Count", fill = "Treatment arm",
       title = "Switching probability distributions under two support scenarios") +
  theme(legend.position = "bottom")
```

::: {.notes}
Under poor support, CKD patients on the active drug switch near-deterministically. The censoring model must extrapolate heavily, and any estimator's performance degrades. This is failure mode #3: right estimand, right estimator, but the data do not support the question.
:::


## Support degrades estimation {.smaller}

```{r}
#| label: tbl-support-results

if (has_sup) {
  sup_display <- sup_res %>%
    mutate(across(where(is.numeric), ~ round(., 3)))
  kable(sup_display, format = "pipe",
        caption = "Estimation results across support scenarios (N = 5,000)")
} else {
  cat("*Support results not cached. Run R/run_cached_analyses.R.*")
}
```

. . .

Under poor support:

- The **censor-at-switch Cox HR diverges** from the naive HR (censoring removes a large, non-random fraction of the active arm)
- **L-TMLE estimates become unstable** as the censoring model extrapolates into poorly supported regions
- Diagnostics — switching probabilities, effective sample sizes, propensity score overlap — should be examined **before** interpreting any estimate

::: {.notes}
The support diagnostic is as important as the estimator choice. A well-aligned estimator with poor support is no better than a misaligned estimator with good support.
:::


## Practical workflow {.smaller}

:::: {.columns}
::: {.column width="60%"}

1. **Define the causal question** — what treatment strategies are being compared? What outcome? Over what time horizon?

2. **Specify intercurrent event handling** — switching, censoring, death: which strategy for each?

3. **Draw the DAG** — identify confounders, mediators, colliders. Determine what is time-varying.

4. **Assess identifiability** — exchangeability, positivity, consistency. State assumptions explicitly.

5. **Choose the estimator** — match it to the estimand, not the reverse.

6. **Diagnose support** — check positivity before trusting any estimate.

7. **Conduct sensitivity analysis** — how robust is the conclusion to violations of key assumptions?
:::

::: {.column width="40%"}

**Common pitfalls:**

- Jumping from question to Cox model without specifying the estimand
- Treating censor-at-switch as "standard" without acknowledging the implied estimand
- Comparing HRs and RDs as though they estimate the same thing
- Interpreting unadjusted KM curves causally
- Ignoring positivity violations
:::
::::


## Which estimand to use? {.smaller}

A practical hierarchy from **most to least feasible** under data limitations, distilled from the manuscript. *Feasibility is not scientific priority — pick the estimand that answers your question, then check support.*

::: {style="font-size: 0.78em;"}

| Rank | Estimand | When it's the right pick | Caveat |
|---|---|---|---|
| 1 | **Treatment-policy** | Default for safety; reflects observed practice; only needs baseline-treatment positivity | Real-world effect including downstream switching — not a pure pharmacological signal |
| 2 | **While-on-treatment** (competing-risks) | Want a descriptive cause-specific 180-day risk; same positivity as TP | Switching halts outcome ascertainment — descriptive quantity, not a counterfactual under sustained treatment |
| 3 | **Composite** | Want to sidestep switching positivity and a single combined endpoint is clinically meaningful | Switching itself counts as an event — inflates the effect (RR 1.46 vs 1.25 for TP) and conflates tolerability with safety |
| 4 | **Principal stratum** | Natural fit for adherence-adjusted, mediation-flavoured, or label-restriction questions | Effect defined on the latent never-switcher subset (~19% of cohort), not the full population; needs cross-world + monotonicity assumptions. The headline 96% coverage is from the *oracle* analysis using cross-world potential outcomes — the observable approximation (restrict to observed non-switchers) achieves only ~50% RD coverage |
| 5 | **No-switch** (hypothetical) | Pure pharmacological effect needed and switching is rare / well-modelled | Most demanding sequential positivity; coverage drops to 87% even at 11% switching — consider a **stochastic intervention** ("reduce switching by 50%") instead when switching exceeds ~40–50% |

:::

. . .

**Three rules that fall out of the simulation:**

- *Specify the estimand before the estimator* — the same data give 1.25–1.46 true RR depending on the strategy.
- *If using Cox, do not interpret the HR as a marginal risk contrast*, and do not assume a Cox-derived RD rescues the analysis under non-proportional hazards.
- *Match the estimator to the estimand and the truth used for validation* — most apparent "estimator failures" we caught were silent estimand mismatches.

::: {.notes}
Anchor each row in a clinical scenario rather than reading verbatim. The 40–50% switching threshold for retiring the no-switch estimand comes from the manuscript's positivity discussion. The no-switch and IPCW interpretation of WOT target the same counterfactual under different identifying assumptions — they are not separate estimands in the population-parameter sense, even though ICH E9(R1) lists them separately. Stochastic interventions (Díaz/Williams/Hoffman LMTP shifts, Rytgaard 2026 calibrated intensity-scaling) sit outside the ICH E9(R1) taxonomy but offer a route when deterministic regimes lack support.
:::


## Conclusions {.smaller}

:::: {.columns}
::: {.column width="60%"}

**Different estimands answer different questions.** Under this DGP the true risk differences ranged from +6.4% (WOT) to +18.3% (composite); under a different scenario the relative ordering can change entirely.

**The ICH E9(R1) strategies are one component of an estimand**, not the full definition — population, time horizon, and outcome must also be specified.

**Cox hazard ratios do not target the estimand.** For treatment-policy, no-switch, and while-on-treatment, the Cox HR 95% CI contained the true marginal HR less than 15% of the time.

**L-TMLE targets the marginal risk contrast directly.** Near-nominal RD and RR coverage (87–96%) across all five estimands once the treatment node is included in the L-TMLE specification.

**Stochastic and realistic interventions are a natural next step** when strict deterministic interventions (e.g., "no one switches") lack data support.

**Diagnostics are not optional.** Identification does not guarantee estimation. Positivity should be assessed iteratively with estimand specification, before interpreting any causal estimate.
:::

::: {.column width="40%"}

**The causal roadmap:**

```{r}
#| label: fig-roadmap-small
#| fig-width: 4.5
#| fig-height: 5

steps_small <- tibble(
  label = c("Question", "DAG", "Identify", "Estimand",
            "Data support", "Estimator", "Sensitivity"),
  y = 7:1,
  loop = c(FALSE, FALSE, FALSE, TRUE, TRUE, TRUE, FALSE)
)

# forward arrows
fwd <- steps_small %>% filter(y > 1) %>%
  mutate(yend = y - 1) %>%
  rowwise() %>% mutate(
    y_start = y - 0.35,
    y_end   = yend + 0.35
  ) %>% ungroup()

# back arrows for the iterative loop (Estimand <-> Data support <-> Estimator)
# nodes: Estimand y=4, Data support y=3, Estimator y=2
back <- tibble(
  y_start = c(3, 2) + 0.20,    # Data support -> Estimand,  Estimator -> Data support
  y_end   = c(4, 3) - 0.20,
  x       = 1.30,
  xend    = 1.30
)

ggplot(steps_small, aes(x = 1, y = y)) +
  # forward (down) arrows
  geom_segment(data = fwd,
               aes(x = 1, xend = 1, y = y_start, yend = y_end),
               arrow = arrow(length = unit(0.15, "cm")),
               color = "grey40") +
  # back (up) arrows showing the iterative loop
  geom_segment(data = back,
               aes(x = x, xend = xend, y = y_start, yend = y_end),
               arrow = arrow(length = unit(0.18, "cm"), ends = "last"),
               color = "tomato", linewidth = 0.9) +
  geom_label(aes(label = label,
                 fill = loop),
             size = 4.5, label.size = 0.5, fontface = "bold") +
  scale_fill_manual(values = c(`FALSE` = "white", `TRUE` = "#FFE9E4"),
                    guide = "none") +
  annotate("text", x = 1.45, y = 3,
           label = "iterate", color = "tomato",
           fontface = "bold.italic", size = 6, hjust = 0) +
  coord_cartesian(xlim = c(0.5, 2.05), ylim = c(0.3, 7.7)) +
  theme_void()
```
:::
::::


## References {.smaller}

::: {style="font-size: 0.65em;"}
- ICH E9(R1) Addendum (2019). Estimands and sensitivity analyses in clinical trials.
- Dang LE, et al. (2023). A causal roadmap for generating high-quality real-world evidence. *J Clin Transl Sci.*
- Petersen ML (2014). Applying a causal road map in settings with time-dependent confounding. *Epidemiology.*
- Gruber S, et al. (2022). Developing a targeted learning-based statistical analysis plan. *Stat Biopharm Res.*
- Gruber S, et al. (2024). Targeted learning: toward a future informed by real-world evidence. *Stat Biopharm Res.*
- Chen RC, et al. (2023). Beyond the Cox hazard ratio. *Clinical Trials.*
- Martinussen T (2022). Causality and the Cox regression model. *Annu Rev Stat Appl.*
- Stensrud MJ, et al. (2019). Limitations of hazard ratios in clinical trials. *Eur Heart J.*
- Rufibach K (2019). Treatment effect quantification for time-to-event endpoints. *Pharm Stat.*
- Williams N, Diaz I (2023). lmtp: Non-parametric causal effects based on modified treatment policies. R package.
- Latimer NR, et al. (2014). Adjusting survival time estimates for treatment switching. *Stat Methods Med Res.*
- Lee SW, et al. (2019). TDF vs ETV on renal function. Systematic review.
- Jung CY, et al. (2022). Kidney function decline with TDF vs ETV. *Int J Infect Dis.*
:::


## {.appendix}

### Appendix: DGP parameters

| Parameter | Value | Description |
|---|---|---|
| `h0` | 5 $\times$ 10$^{-3}$ | Baseline hazard |
| `HR_early` | 2.0 | TDF vs ETV, first 90 days |
| `HR_late` | 0.90 | TDF vs ETV, after 90 days |
| `lambda_sw0` | 1.0 $\times$ 10$^{-3}$ | Baseline switching hazard |
| `gamma_A` | 0.80 | log-HR for treatment on switching |
| `gamma_ckd` | 0.60 | log-HR for CKD on switching |
| CKD prevalence | 8% | Key confounder |
| N (simulation) | 10,000 | Per iteration |
| Time bins | 14-day (biweekly) | For L-TMLE discretisation |

### Appendix: estimand–estimator alignment table

| Estimand | Cox naive | Cox censor-at-switch | L-TMLE |
|---|---|---|---|
| Treatment-policy | Conditional HR (not marginal RD) | Wrong estimand (censors at switch) | Baseline trt; marginal RD under static intervention |
| No-switch | Includes post-switch outcomes | Assumes non-informative censoring | Time-varying trt; holds constant = no-switch |
| While-on-treatment | Includes post-switch person-time | Biased if switching informative | Baseline trt; switch-censoring in C columns |
| Composite | HR for composite endpoint | Same | Baseline trt; composite in Y columns |
| Principal stratum | Not targeted | Not targeted | Baseline trt on oracle never-switcher subset |
